\section{Basic Definitions}


\begin{defi}
 A \emph{$k$-Lie algebra} is a vector space $\g$ (over some field $k$) together with a $k$-bilinear map
 \[
  [\cdot, \cdot] \colon \g \times \g \to \g
 \]
 satisfying the following:
 \begin{enumerate}
  \item
   $[\cdot, \cdot]$ is \emph{alternating}, i.e.\ $[x,x] = 0$ for every $x \in \g$.
  \item
   The \emph{Jacobi identity}
   \[
    [x,[y,z]] + [y,[z,x]] + [z,[x,y]] = 0
    \quad
    \text{for all $x,y,z \in \g$}.
   \]
 \end{enumerate}
 $[\cdot,\cdot]$ is called a \emph{Lie bracket}.
\end{defi}


\begin{rem}
 $[\cdot, \cdot]$ is antisymmetric, i.e.\ $[y,x] = -[x,y]$ for all $x,y \in \g$, because
 \[
  0 = [x+y, x+y] = [x,x] + [x,y] + [y,x] + [y,y] = [x,y] + [y,x].
 \]
\end{rem}


\begin{defi}
 Let $A$ be a $k$-algebra. A \emph{derivation of $A$} is a $k$-linear map $d \colon A \to A$ such that
 \[
  d(ab) = d(a)b + ad(b) \quad \text{for al l$a,b \in A$}.
 \]
 We set
 \[
  \Der(A) \coloneqq \{d \colon A \to A \mid \text{$d$ is a derivation of $A$} \}.
 \]
\end{defi}


\begin{rem}
 $\Der(A)$ is clearly a $k$-vector space.
\end{rem}


\begin{expls}
 \begin{enumerate}
  \item
   Any vector space $V$ becomes a Lie algebra via
   \[
    [x,y] = 0 \quad \text{for all $x,y \in V$}.
   \]
  \item
   Any associative $k$-algebra $A$ becomes a Lie algebra via
   \[
    [a,b] = ab-ba \quad \text{for all $a,b \in A$}.
   \]
   It is clear that $[\cdot, \cdot]$ is alternating and the Jacobi identity can be verified by some easy calculation.
   
   In particular $\Mat_n(k)$ is a Lie algebra via
   \[
    [A,B] = AB-BA \quad \text{for all $A,B \in \Mat_n(k)$}.
   \]
   This is called the \emph{general linear Lie algebra} and is denoted by $\gl_n(k)$ or $\gl(n,k)$.
   
   More generally for any $k$-vector space the space $\End_k(V)$ becomes a Lie algebra via
   \[
    [\varphi_1, \varphi_2] \coloneqq \varphi_1 \circ \varphi_2 - \varphi_2 \circ \varphi_1
    \quad
    \text{for all $\varphi_1, \varphi_2 \in \End_k$}.
   \]
   This is called the \emph{general linear Lie algebra for $V$} and is denoted by $\gl(V)$.
   
   A Lie algebra $\g$ is called linear if $\g \subseteq \gl(V)$ for some finite dimensional vector space $V$.
  \item
   Let $A$ be a $k$-algebra. $\Der(A)$ is a Lie algebra via
   \[
    [d, d'] \coloneqq d \circ d' - d' \circ d
    \quad 
    \text{for all $d, d' \in \Der(A)$}.
   \]
   Is is an easy calculation to show that $[d,d']$ is again a derivation. Notice that the Jacobi identity for $\Der(A)$ follows from the Jacobi identity for $\gl(A)$.
 \end{enumerate}
\end{expls}


We will now look at how to construct new Lie algebras from old ones.


\begin{defi}
 Let $\g_1$ and $\g_2$ be Lie algebras over the same field $k$. Then the \emph{product} of $\g_1$ and $\g_2$ is defined as the $k$-vector space $\g_1 \times \g_2$ together with the Lie bracket
 \[
  [(x_1, y_1), (x_2, y_2)]
  = ([x_1, x_2], [y_1, y_2])
  \quad
  \text{for all $(x_1, y_1), (x_2, y_2) \in \g_1 \times \g_2$}.
 \]
\end{defi}


Let $\g$ be a Lie algebra over $k$ and $A$ an associative, commutative $k$-algebra. Then $\g \otimes_k A$ is a Lie algebra via
\[
 [x \otimes a, y \otimes b] = [x,y] \otimes (ab)
 \quad
 \text{for all $x,y \in \g$ and $a,b \in A$}.
\]


\begin{defi}
 Let $\g$ be a Lie algebra and $A = k[t,t^{-1}]$ be the algebra of Laurent polynomials over $k$. Then
 \[
  \Loop(\g) \coloneqq \g \otimes_k A
 \]
 with the Lie bracket as above is called the \emph{loop (Lie) algebra} of $\g$.
\end{defi}


Another example of construction new Lie algebras from old ones are \emph{central extensions}: Let $\g$ be any $k$-Lie algebra.
\[
 \tilde{\g}
 \coloneqq \g \otimes k
 = \{x + \lambda c \mid x \in \g, \lambda \in k\},
\]
where we understand $c$ as a formal variable. Suppose that $\kappa \colon \g \times \g \to k$ is a $k$-bilinear map satisfying the following properties:
\begin{enumerate}
 \item
  $\kappa$ is antisymmetric, i.e.\ $\kappa(x,y) = -\kappa(y,x)$ for all $x,y \in \g$.
 \item
  The $2$-cocycle condition
  \[
   \kappa([x,y],z) + \kappa([y,z],x) + \kappa([z,x],y) = 0
   \quad
   \text{for all $x,y,z \in \g$}.
  \]
\end{enumerate}
Then $\tilde{\g}$ becomes a Lie algebra via
\[
 [x + \lambda c, y + \mu c] \coloneqq [x,y] + \kappa(x,y) \lambda \mu c.
\]
Note that $c$ is central in $\tilde{\g}$ in the sense that $[x,c] = 0$ for all $x \in \g$.


\begin{expl}
 Let $\g = \gl_n(k)$. We define a symmetric bilinear form on $\g$ via
 \[
  (A,B)_{\tr} = \tr(AB).
 \]
 We define a bilinear form
 \[
  \Loop(\g) \times \Loop(\g) \to k[t,t^{-1}],
  (x \otimes p, y \otimes q) \mapsto (x,y)_{\tr}\ pq
 \]
 We now get a $2$-cocycle $\kappa \colon \Loop(\g) \times \Loop(\g) \to k$ via
 \[
  \kappa(a,b) \coloneqq \mathrm{Res}\left(\frac{\partial a}{\partial t}, b\right).
 \]
 $\kappa$ is also antisymmetric: Let $a = x \otimes t^i$ and $b = y \otimes t^j$ with $x,y \in \g$ and $i,j \in \Z$. Then
 \begin{align*}
  \kappa(x \otimes t^i, y \otimes t^j)
  = \mathrm{Res}(i x \otimes t^{i-1}, y \otimes t^j)
  &= \mathrm{Res}(i t^{i+j-1} (x,y)_{\tr}) \\
  &=
  \begin{cases}
   i (x,y)_{\tr} & \text{if $i+j = 0$}, \\
               0 & \text{otherwise}.
  \end{cases}
 \end{align*}
 In the same way we find that
 \[
  \kappa(y \otimes t^j, x \otimes t^i) =
  \begin{cases}
   j (x,y)_{\tr} & \text{if $i+j = 0$}, \\
               0 & \text{otherwise}.
  \end{cases}
 \]
 Since $(\cdot,\cdot)_{\tr}$ is symmetric we find that
 \begin{align*}
  \kappa(x \otimes t^i, y \otimes t^j)
  &=
  \begin{cases}
   i (x,y)_{\tr} & \text{if $i+j = 0$}, \\
               0 & \text{otherwise},
  \end{cases} \\
  &=
  \begin{cases}
   -j (x,y)_{\tr} & \text{if $i+j = 0$}, \\
                0 & \text{otherwise},
  \end{cases} \\
  &=
  -\kappa(y \otimes t^j, x \otimes t^i).
 \end{align*}
\end{expl}


As for all algebraic structures morphisms are of interest.


\begin{defi}
 Given $k$-Lie algebras $\g_1$ and $\g_2$ a homomorphism of Lie algebras $\g_1 \to \g_2$ is a $k$-linear map $f \colon \g_1 \to \g_2$ such that
 \[
  f([x,y]) =[f(x),f(y)] \quad \text{for all $x,y \in \g_1$}.
 \]
\end{defi}


\begin{expls}
 \begin{enumerate}
  \item
   For any Lie algebra $\g$ the identity $\id_\g \colon \g \to \g$ is a Lie algebra homomorphism.
  \item
   Given Lie algebras $\g_1$, $\g_2$ and $\g_3$ and Lie algebra homomorphisms $f_1 \colon \g_1 \to \g_2$ and $f_2 \colon \g_2 \to \g_3$ the composition $f_2 \circ f_1 \colon \g_1 \to \g_3$ is also a homomorphism of Lie algebras.
 \end{enumerate}
\end{expls}


\begin{rem}
 We find that we have a category of $k$-Lie algebras.
\end{rem}


\begin{defi}
 Let $\g$ be a $k$-Lie algebra. A representation of $\g$ is a $k$-vector space $V$ together with a homomorphism of Lie algebras $\rho \colon \g \to \gl(V)$.
\end{defi}


\begin{rem}
  Equivalently a representation of $\g$ is a $k$-vector space $V$ together with a $k$-bilinear map $\g \times V \to V, (x,v) \mapsto x.v$ such that
 \[
  x.y.v - y.x.v = [x,y].v \quad \text{for all $x,y \in \g$ and $v \in V$}.
 \]
\end{rem}


\begin{defi}
 Let $\g$ be a Lie algebra. Then
 \[
  \ad \colon \g \to \gl(\g), x \mapsto [x,\cdot]
 \]
 is called the \emph{adjoint representation} of $\g$.
\end{defi}


\begin{rem}
 That $\ad$ is a homomorphism of Lie algebras is equivalent to
 \[
  \ad([x,y])(z) = [\ad(x),\ad(y)](z) \quad \text{for all $x,y,z \in \g$}.
 \]
 Because
 \begin{gather*}
  \ad([x,y])(z) = [[x,y],z] = -[z,[x,y]]
 \shortintertext{and}
  \begin{aligned}
   [\ad(x),\ad(y)](z)
   &= (\ad(x) \circ \ad(y))(z) - (\ad(y) \circ \ad(x))(z) \\
   &= [x,[y,z]] - [y,[x,z]]
   = [x,[y,z]] + [y,[z,x]]
  \end{aligned}
 \end{gather*}
 this is equivalent to the Jacobi identity.
\end{rem}


\begin{defi}
 Let $\g$ be a $k$-Lie algebra. A \emph{Lie subalgebra} of a $\g$ is a $k$-linear subspace $L \subseteq \g$ such that
 \[
  [x,y] \in L \quad \text{for all $x,y \in L$}.
 \]
 An \emph{ideal} inside $\g$ is a $k$-linear subspace $I \subseteq \g$ such that
 \[
  [x,y] \in I \quad \text{for all $x \in \g$ and $y \in I$}.
 \]
 We denote ideals by $I \subideal \g$.
\end{defi}


For a Lie algebra $\g$ and a subalgebra $\h \subseteq \g$ it is clear that $\h$ becomes a Lie algebra by restricting the Lie bracket of $\g$ to $\h$. The inclusion $\h \inc \g$  is then clearly a homomorphism of Lie algebras.

Also notice that every ideal inside $\g$ is also a subalgebra of $\g$.


\begin{expl}
 Let $\g = \gl_n(k)$. Then
 \[
  \Sl_n(k) = \{A \in \g \mid \tr A = 0\}
 \]
 is a subalgebra. Even more
 \[
  \Sl_n(k) = [\g,\g].
 \]
 To see this first notice that on the one hand $A, B \in \g$
 \[
  \tr [A,B]
  = \tr(AB-BA)
  = \tr(AB) - \tr(BA)
  = \tr(AB) - \tr(AB)
  = 0.
 \]
 Therefore $[\g,\g] \subseteq \Sl_n(k)$. On the other hand notice that $\Sl_n(k)$ has a basis given by the elementary matrices $e_{ij}$ with $1 \leq i \neq j \leq n$ and $e_{11} - e_{ii}$ with $1 < i \leq n$. Each of these matrices is given as a commutator, namely $e_{ij} = [e_{ii},e_{ij}]$ for $1 \leq i \neq j \leq n$ and $e_{11} - e_{ii} = [e_{1i},e_{i1}]$ for $1 < i \leq n$. Therefore $\Sl_n(k) \subseteq [\g,\g]$.
\end{expl}



We have similar properties for ideals inside of Lie algebras als for ideals inside of a ring: For a collection of ideals $I_\lambda$, $\lambda \in \Lambda$, their intersection $\bigcap_{\lambda \in \Lambda} I_\lambda$ and their sum $\sum_{\lambda \in \Lambda} I_\lambda$ are also ideals inside $\g$. For ideals $I, J \subideal \g$ the subspace $[I,J]$ is also an ideal inside $\g$. (For any two subsets $X, Y \subseteq \g$ we denote by $[X,Y]$ the linear subspace generated by the commutators $[x,y]$ with $x \in X$, $y \in Y$.) To see that $[I,J]$ is an ideal notice that for any $x \in I$, $y \in J$ and $z \in \g$
\[
 [z,[x,y]] = -[x,\underbrace{[y,z]}_{\in J}]-[y,\underbrace{[z,x]}_{\in I}].
\]


\begin{rem}
 If $\g$ is a Lie algebra and $I \subideal \g$ then the quotient vector space $\g/I$ is also a Lie algebra via
 \[
  [x+I, y+I] = [x,y] + I.
 \]
 It is easy to see that this bilinear map is well-defined. The properties for the Lie bracket on $\g/I$ follow from the ones for the Lie bracket on $\g$.
 
 From the definition of the Lie bracket on $\g/I$ it is clear that the canonical projection $\pi \colon \g \to \g/I$ is a homomorphism of Lie algebras.
\end{rem}


We have the usual theorems about homomorphisms and ideals.


\begin{prop}
 Let $\g_1$ and $\g_2$ be Lie algebras and $f \colon \g_1 \to \g_2$ a homomorphism of Lie algebras.
 \begin{enumerate}
  \item
   $\ker f \subideal \g_1$ is an ideal.
  \item
   $\im f \subseteq \g_2$ is a Lie subalgebra.
  \item
   If $I \subideal \g_1$ is an ideal with $\ker f \subseteq I$ then there exists a unique homomorphism of Lie algebras $\tilde{f} \colon \g_1/I \to \g_2$ with $f = \tilde{f} \circ \pi$ where $\pi \colon \g_1 \to \g_1/I$ is the canonical projection.
   \tikzsetnextfilename{universal_property_of_quotient_of_lie_algebras}
   \begin{center}
    \begin{tikzpicture}[node distance = 2cm]
     \node (g1) {$\g_1$};
     \node (gI) [below of = g1] {$\g_1/I$};
     \node (g2) [right of = gI] {$\g_2$};
     \draw[->] (g1) to node[above right] {$f$} (g2);
     \draw[->] (g1) to node[left] {$\pi$} (gI);
     \draw[->] (gI) to node[below] {$\exists! \tilde{f}$} (g2);
    \end{tikzpicture}
   \end{center}
  \item
   If $I, J \subideal \g$ are subideals with $I \subseteq J$ then the map
   \[
    (\g/I)/(J/I) \to \g/I, (x+I)+(J/I) \mapsto x+I
   \]
   is a natural isomorphism.
  \item
   If $I, J \subideal \g$ are subideals then there is a natural isomorphism
   \begin{gather*}
    (I + J)/J \to I/(I \cap J)
   \shortintertext{defined by}
    (x+J)+I \mapsto x + (I \cap J) \quad \text{ for every $x \in I$}
   \end{gather*}
   is a natural isomorphism.
 \end{enumerate}
\end{prop}


\begin{defi}
 Let $\g$ be a Lie algebra. The \emph{center} of $\g$ is
 \[
  Z(\g)
  \coloneqq \{x \in \g \mid \text{$[x,y] = 0$ for every $y \in \g$}\}
  = \ker \ad.
 \]
 
 $\g$ is called \emph{abelian} if $[\g,\g] = 0$.
\end{defi}


Clearly $\g$ is abelian if and only if $Z(\g) = \g$. Notice that $[\g,\g]$ and $Z(\g)$ are ideals inside $\g$.


\begin{defi}
 A Lie algebra $\g$ is \emph{simple} if $0$ and $\g$ are the only ideals inside $\g$ and $\g$ is not abelian.
\end{defi}


\begin{lem}
 Let $\g$ be a simple Lie algebra. Then $[\g,\g] = \g$ and $Z(\g)=0$.
\end{lem}
\begin{proof}
 Because $\g$ is simple it is not abelian. Therefore $[\g,\g] \neq 0$ and $Z(\g) \neq \g$. Since $[\g,\g]$ and $Z(\g)$ are ideals inside $\g$ it follows that $[\g,\g] = \g$ and $Z(\g) = 0$.
\end{proof}


\begin{cor}
 Let $\g$ be simple. Then $\ad \colon \g \to \gl(\g)$ is injective. In particular $\g$ is linear.
\end{cor}
\begin{proof}
 This directly follows from $\ker \ad = Z(\g) = 0$.
\end{proof}


\begin{thrm}[Ado]
 Any finite dimensional Lie algebra is linear.
\end{thrm}


A proof of Ado’s theorem will (maybe) be given later.


\begin{rem}
 For a Lie algebra $\g$ the ideal $[\g,\g]$ is the minimal ideal inside $\g$ such that $\g/[\g,\g]$ is abelian. Furthermore given any abelian Lie algebra $\h$ any homomorphism of Lie algebras $\g \to \h$ factorizes through a unique homomorphism of Lie algebras $\g/[\g,\g] \to \h$.
 \begin{center}
  \tikzsetnextfilename{universal_property_of_abelization}
  \begin{tikzpicture}[node distance = 2cm]
   \node (gab) {$\g/[\g,\g]$};
   \node (g) [above left of = gab] {$\g$};
   \node (h) [above right of = gab] {$\h$};
   \draw[->] (g) to (h);
   \draw[->] (g) to (gab);
   \draw[->] (gab) to node[below right] {$\exists!$} (h);
  \end{tikzpicture}
 \end{center}
\end{rem}


\begin{expls}
 \begin{enumerate}
  \item
   Since $[\gl_n(k),\gl_n(k)] = \Sl_n(k) \neq \gl_n(k)$ we find that $\gl_n(k)$ is not simple.
  \item
   Let $\g = \Sl_2(k)$. Then $\g$ is simple if and only if $\kchar k \neq 2$. To see this consider the basis
   \[
    e = \begin{pmatrix}0 & 1 \\ 0 & 0\end{pmatrix}, h = \vect{1 & 0 \\ 0 & -1}, f = \vect{0 & 0 \\ 1 & 0}.
   \]
   of $\Sl_2(k)$. Then
   \[
    [e,h] = -2e, [e,f] = h, [h,f] = -2f.
   \]
   If $\kchar k = 2$ then $h$ spans a $1$-dimensional ideal, thus $\Sl_2(k)$ is not simple. Suppose that $\kchar k \neq 2$ and let $I \subseteq \Sl_2(k)$ be an ideal with $I \neq 0$. It is clear that if $I$ contains one of the basis vectors $e$, $h$ or $f$ it follows that $I = \Sl_2(k)$. Let $x \in I$ with $x \neq 0$ and write $x = \alpha e + \beta h + \gamma f$. Then
   \[
    [e,x] = -2 \beta e + \gamma h \quad \text{and} \quad [e,[e,x]] = -2 \gamma e.
   \]
   Since $\gamma = 0$ or $\gamma \neq 0$ we find that $e \in I$.
 \end{enumerate}
\end{expls}


\begin{rem}
 $\Sl_n(\C)$ is simple for all $n \geq 2$.
\end{rem}





\section{Nilpotent and solvable Lie algebras}

\begin{defi}
 Let $A$ be a $k$-algebra. An element $a \in A$ is called \emph{nilpotent} if $a^n = 0$ for some $n \geq 1$. Given a Lie algebra $\g$ an element $x \in \g$ is called \emph{$\ad$-nilpotent} if $\ad(x) \in \End_k(\g)$ is nilpotent.
\end{defi}


\begin{defi}
 Let $g$ be a Lie algebra. Define $\g^0 \coloneqq \g$ and $\g^{i+1} \coloneqq [\g,\g^i]$ for all $i \geq 0$. Then
 \[
  \g = \g^0 \supseteq \g^1 \supseteq \g^2 \supseteq \dots
 \]
 is called the \emph{central series} of $\g$. Also define $\g^{(0)} \coloneqq \g$ and $\g^{(i+1)} \coloneqq [\g^{(i)},g^{(i)}]$ for all $i \geq 0$. Then
 \[
  \g^{(0)} \supseteq \g^{(1)} \supseteq \g^{(2)} \supseteq \dots
 \]
 is called the \emph{derived series} of $\g$. $\g$ is called \emph{nilpotent} if $\g^i = 0$ for some $i$ and \emph{solvable} if $g^{(i)} = 0$ for some $i$.
\end{defi}


It is clear that every nilpotent Lie algebra is also solvable.


\begin{expls}
 \begin{enumerate}
  \item
   Because $[\gl_n(\C),\gl_n(\C)] = \Sl_n(\C)$ and $[\Sl_n(\C),\Sl_n(\C)] = \Sl_n(\C)$ for $n \geq 2$ (since $\Sl_n(\C)$ is simple) $\gl_n(\C)$ is not simple.
  \item
   If $\g$ is abelian then $\g$ is nilpotent.
  \item
   A Heisenberge Lie algebras consists of a real vector space with basis $P_1, \dotsc, P_n$, $Q_1, \dotsc, Q_n$, $C$ together with the Lie bracket satisfying the following conditions:
   \[
    [P_i, P_j] = [Q_i, Q_j] = [P_i, C] = [Q_i, C] = 0
    \quad \text{and} \quad
    [P_i, Q_j] = \delta_{ij} C.
   \]
   This defines a nilpotent Lie algebra.
 \end{enumerate}
\end{expls}


\begin{prop}
 Let $\g$ be a Lie algebra.
 \begin{enumerate}
  \item
   If $\h$ is a Lie algebra and $f \colon \g \to \h$ aLie algebras homomorphism then $f(\g)^i = f(\g^i)$ and $f(\g)^{(i)} = f(\g^{(i)})$ for all $i \geq 0$.
  \item
   If $\g$ is nilpotent (resp.\ solvable) then any Lie subalgebra $\h \subseteq \g$ and any quotient of $\g$ (by an ideal) is nilpotent (resp.\ nilpotent).
  \item
   If $I \subideal \g$ with $I \subseteq Z(\g)$ and $\g/I$ is nilpotent then $\g$ is nilpotent.
  \item
   If $\g \neq 0$ is nilpotent then $Z(\g) \neq 0$.
  \item
   If $\g$ is nilpotent and $x \in \g$ then $x$ is $\ad$-nilpotent.
  \item
   If $I \subideal \g$ then $I^i$ and $I^{(i)}$ are ideals inside $\g$ for all $i \geq 0$.
 \end{enumerate}
\end{prop}
\begin{proof}
 \begin{enumerate}
  \item
   It suficies to show that for any two subsets $X, Y \subseteq \g$
   \[
    f([X,Y])= [f(X),f(Y)]
   \]
   the statement then follows inductively. It holds because $f$ is a Lie algebra homomorphism and therefore
   \begin{align*}
    f([X,Y])
    &= f(\vspan_k \{[x, y] \mid x \in X, y \in Y\}) \\
    &= \vspan_k \{f([x, y]) \mid x \in X, y \in Y\} \\
    &= \vspan_k \{[f(x),f(y)] \mid x \in X, y \in Y\} \\
    &= \vspan_k \{[x',y'] \mid x' \in f(X), y' \in f(Y)\} \\
    &= [f(X),f(Y)].
   \end{align*}
  \item
   The statement about subalgebras is clear since $\h^i \subseteq \g^i$ and $\h^{(i)} \subseteq \g^{(i)}$ for all $i$. The statement about quotient follow using the canonical projection $\pi \colon \g/I$ where $I \subideal \g$. Because $\pi$ is a Lie algebra homomorphism we have
   \[
    (\g/I)^i = \pi(\g)^i = \pi(\g^i) = 0
   \]
   for $i$ big enough. For solvable $\g$ the corresponding statements follow in the same way.
  \item
   Let $\pi \colon \g \to \g/I$ be the canonical projection. Because $\g/I$ is nilpotent there exists some $i \geq 0$ with $(\g/I)^i = 0$ and therefore
   \[
    0 = (\g/I)^i = \pi(\g)^i = \pi(\g^i).
   \]
   Thus $\g^i \subseteq I \subseteq Z(G)$ and hence $\g^{i+1} = 0$.
  \item
   Let $i \geq 1$ be minimal such that $\g^{i-1} \neq 0$ but $\g^i = 0$. Then $\g^{i-1} \subseteq Z(\g)$ and therefore $Z(\g) \neq 0$.
  \item
   Since $\g$ is nilpotent $\g^i = 0$ for some $i \geq 0$. Then
   \[
    (\ad(x))^i(\g) \subseteq \g^i = 0,
   \]
   so $(\ad(x))^i = 0$.
  \item
   This follows inductively by using that $[I,J]$ is an ideal inside $\g$ for any $I,J \subideal \g$.
 \end{enumerate}
\end{proof}













































