\section{Universal Enveloping Algebra}





\subsection{Definition}


\begin{definition}
	Let~$\glie$ be a~\liealgebra{$\kf$}.
	A \defemph{universal enveloping algebra}\index{universal enveloping algebra} of the Lie~algebra~$\glie$ is a~\algebra{$\kf$}~$\Univ(\glie)$\glsadd{universal enveloping algebra} together with a homomorphism of Lie~algebras~$\iota$ from~$\glie$ to~$\Univ(\glie)$ such that the following universal property\index{universal property! of the universal enveloping algebra} holds:
	For every~{\algebra{$\kf$}}~$A$ and every homomorphism of Lie~algebras~$\varphi$ from~$\glie$ to~$A$ there exists a unique homomorphism of~\algebras{$\kf$}~$\Phi$  from~$\Univ(\glie)$ to~$A$ that makes the triangular diagram
	\[
		\begin{tikzcd}
			\glie
			\arrow{r}[above]{\varphi}
			\arrow{d}[left]{\iota}
			&
			A
			\\
			\Univ(\glie)
			\arrow[dashed]{ur}[below right]{\Phi}
			&
			{}
		\end{tikzcd}
	\]
	commute, i.e.\ such that~$\varphi = \Phi \circ \iota$.
	In other words, every homomorphism of Lie~algebras from~$\glie$ to~$A$ extends along~$\iota$ uniquely to a homomorphism of algebras from~$\Univ(\glie)$ to~$A$.
	The homomorphism~$\iota$ is the \defemph{canonical homomorphism}\index{canonical homomorphism} from~$\glie$ to~$\Univ(\glie)$.
\end{definition}


\begin{remark}[Uniqueness of universal enveloping algebras]
	\label{uniqueness of universal enveloping algebras}
	Let~$\glie$ be a Lie algebra and suppose that~$(\Univ(\glie)_1, \iota_1)$ and~$(\Univ(\glie)_2, \iota_2)$ are two~{\uas} of~$\glie$.
	Then there exist unique homomorphisms of algebras~$\Phi$ from~$\Univ(\glie)_1$ to~$\Univ(\glie)_2$ and~$\Psi$ from~$\Univ(\glie)_2$ to~$\Univ(\glie)_1$ that make the triangular diagrams
	\[
		\begin{tikzcd}[column sep = small]
			{}
			&
			\glie
			\arrow{dl}[above left]{\iota_1}
			\arrow{dr}[above right]{\iota_2}
			&
			{}
			\\
			\Univ(\glie)_1
			\arrow[dashed]{rr}[below]{\Phi}
			&
			{}
			&
			\Univ(\glie)_2
		\end{tikzcd}
		\qquad\text{and}\qquad
		\begin{tikzcd}[column sep = small]
			{}
			&
			\glie
			\arrow{dl}[above left]{\iota_2}
			\arrow{dr}[above right]{\iota_1}
			&
			{}
			\\
			\Univ(\glie)_2
			\arrow[dashed]{rr}[below]{\Psi}
			&
			{}
			&
			\Univ(\glie)_1
		\end{tikzcd}
	\]
	commute.
	It follows that the composites~$\Psi \circ \Phi$ and~$\Phi \circ \Psi$ make the triangle diagrams
	\[
		\begin{tikzcd}[column sep = small]
			{}
			&
			\glie
			\arrow{dl}[above left]{\iota_1}
			\arrow{dr}[above right]{\iota_1}
			&
			{}
			\\
			\Univ(\glie)_1
			\arrow[dashed]{rr}[below]{\Psi \circ \Phi}
			&
			{}
			&
			\Univ(\glie)_1
		\end{tikzcd}
		\qquad\text{and}\qquad
		\begin{tikzcd}[column sep = small]
			{}
			&
			\glie
			\arrow{dl}[above left]{\iota_2}
			\arrow{dr}[above right]{\iota_2}
			&
			{}
			\\
			\Univ(\glie)_2
			\arrow[dashed]{rr}[below]{\Phi \circ \Psi}
			&
			{}
			&
			\Univ(\glie)_2
		\end{tikzcd}
	\]
	commute.

	The algebra homomorphisms~$\Phi \circ \Psi$ and~$\Psi \circ \Phi$ are unique with this property by the universal properties of the {\uas}~$(\Univ(\glie)_1, \iota_1)$ and~$(\Univ(\glie)_2, \iota_2)$.
	But the identities~$\id_{\Univ(\glie)_1}$ and~$\id_{\Univ(\glie)_2}$ also makes these diagrams commute.
	We thus find that the composite~$\Psi \circ \Phi$ equals~$\id_{\Univ(\glie)_1}$ and that the composite~$\Phi \circ \Psi$ equals~$\id_{\Univ(\glie)_2}$.
	The homomorphisms~$\Phi$ and~$\Psi$ are therefore mutually inverse isomorphisms of algebras.
	
	This shows that a universal enveloping algebra of~$\glie$ is unique up to unique isomorphism.
	We will therefore talk about \emph{the} universal enveloping algebra of~$\glie$.
	We will often also surpress the homorphism of Lie~algebra~$\iota$ from~$\glie$ to~$\Univ(\glie)$ from our notation.
\end{remark}


% \begin{remark}
%   One can also formulate the above argument is a more categorical way:
%   Consider the category~$\catC$ where
%   \begin{itemize}
%     \item
%       objects of~$\catC$ is a pairs~$(A, i)$ consisting of a~{\algebra{$\kf$}}~$A$ and a Lie~algebra homomorphism~$i \colon \glie \to A$,
%     \item
%       a morphism~$\phi \colon (A, i) \to (B, j)$ is an algebra homomorphism~$\phi \colon A \to B$ that makes the triangular diagram
%       \[
%         \begin{tikzcd}[column sep = small]
%         {}
%         &
%         \glie
%         \arrow{dl}[above left]{i}
%         \arrow{dr}[above right]{j}
%         &
%         {}
%         \\
%         A
%         \arrow[dashed]{rr}[below]{\phi}
%         &
%         {}
%         &
%         B
%       \end{tikzcd}
%     \]
%       commute, and
%     \item
%       the composition of two morphisms is just their usual set-theoretic composition.
%   \end{itemize}
%   A universal enveloping algebra of~$\glie$ is nothing but an inital object in this category~$\catC$.
%   The argumentation from \cref{uniqueness of universal enveloping algebras} is then the usual argument for the uniqueness of inital objects up to unique isomorphism.
% \end{remark}



\subsection{Construction}


\begin{fluff}
	Let~$\glie$ be a Lie~algebra.
	We will in the following show that the universal enveloping algebra of~$\glie$ exists.
	To figure out the construction of~$\Tensor(\glie)$ we will use the universal property of~$\Univ(\glie)$ to explain that we should be able to construct~$\Univ(\glie)$ as a certain quotient algebra of the tensor algebra~$\Tensor(\glie)$.
	We then show that this quotient does indeed have the correct universal property.

	Suppose that~$\glie$ admits a universal enveloping algebra~$\Univ(\glie)$ and let~$\iota$ be the canonical homomorphism of Lie~algebras from~$\glie$ to~$\Univ(\glie)$.
	Let us first observe that the algebra~$\Univ(\glie)$ is generated by the image of~$\iota$.

	Indeed, let~$U$ be the subalgebra of~$\Univ(\glie)$ that is generated by the image of~$\iota$, and let~$\iota'$ be the restriction of~$\iota$ to a homomorphism of Lie~algebras from~$\glie$ to~$U$.
	For every~{\algebra{$\kf$}}~$A$ and every Lie~algebra homomorphism~$\varphi$ from~$\glie$ to~$A$ the induced homomorphism of algebras~$\Phi$ from~$\Univ(\glie)$ to~$A$ restricts to an homomorphism of algebras~$\Phi'$ from~$U$ to~$A$.
	This homomorphism~$\Phi'$ makes the triangular diagram
	\[
		\begin{tikzcd}
			\glie
			\arrow{r}[above]{\varphi}
			\arrow{d}[left]{\iota'}
			&
			A
			\\
			U
			\arrow[dashed]{ur}[below right]{\Phi'}
			&
			{}
		\end{tikzcd}
	\]
	commute.
	The homomorphism~$\Phi'$ is unique with this property because the algebra~$U$ is generated by the image of~$\iota'$, and the composite~$\Phi \circ \iota'$ equals the predetermined homomorphism of Lie~algebras~$\varphi$.
	This shows that the algebra~$U$ together with the homomorphism of Lie~algebras~$\iota'$ is again a universal enveloping algebra for~$\glie$.
	
	It follows from the uniqueness of the universal enveloping algebra of~$\glie$, as discussed in \cref{uniqueness of universal enveloping algebras}, that there exists a unique homomorphism of algebras~$\Iota$ from~$U$ to~$\Univ(\glie)$ that makes the triangular diagram
	\[
		\begin{tikzcd}[column sep = small]
			{}
			&
			\glie
			\arrow{dl}[above left]{\iota'}
			\arrow{dr}[above right]{\iota}
			&
			{}
			\\
			U
			\arrow[dashed]{rr}[below]{\Iota}
			&
			{}
			&
			\Univ(\glie)
		\end{tikzcd}
	\]
	commute, and that this homomorphism is already an isomorphism.%
	\footnote{The letter~$\Iota$ that we use here is the uppercase version of the greek letter iota.}
	The homomorphism~$\Iota$ is the inclusion map from~$U$ to~$\Univ(\glie)$ because this is a homomorphism of algebras from~$U$ to~$\Univ(\glie)$ that makes the above triangular diagram commute.
	We have thus found that the inclusion map from~$U$ to~$\Univ(\glie)$ is an isomorphism of algebras, whence~$U$ equals~$\Univ(\glie)$.

	We now apply the universal property of the tensor algebra~$\Tensor(\glie)$ to the linear map~$\iota$.
	We find that there exists a unique homomorphism of algebras~$\Phi'$ from~$\Tensor(\glie)$ to~$\Univ(\glie)$ that makes the triangular diagram
	\[
		\begin{tikzcd}[column sep = small]
			{}
			&
			\glie
			\arrow{dl}
			\arrow{dr}[above right]{\iota}
			&
			{}
			\\
			\Tensor(\glie)
			\arrow[dashed]{rr}[below]{\Phi'}
			&
			{}
			&
			\Univ(\glie)
		\end{tikzcd}
	\]
	commute.
	The homomorphism~$\Phi'$ is surjective because~$\Univ(\glie)$ is generated by the image of~$\iota$ as an algebra.
	It follows that~$\Phi'$ induces an isomorphism of algebras
	\[
		\Phi
		\colon
		\Tensor(\glie) / I
		\to
		\Univ(\glie)
	\]
	where the ideal~$I$ is the kernel of~$\Phi'$.
	This isomorphism makes the resulting diagram
	 \[
		\begin{tikzcd}[column sep = small]
			{}
			&
			\glie
			\arrow[bend right]{dl}
			\arrow[bend left]{ddr}[above right]{\iota}
			&
			{}
			\\
			\Tensor(\glie)
			\arrow[bend left]{drr}[below left]{\Phi'}
			\arrow{d}
			&
			{}
			&
			{}
			\\
			\Tensor(\glie)/I
			\arrow[dashed]{rr}[below]{\Phi}
			&
			{}
			&
			\Univ(\glie)
		\end{tikzcd}
	\]
	commute.
 
	Let~$A$ be another~\algebra{$\kf$}.
	Every linear map~$g$ from~$\glie$ to~$A$ factors through a homomorphism of algebras~$\Psi'$ from~$\Tensor(\glie)$ to~$A$.
	It follows thanks to the above isomorphism~$\Phi$ between~$\Tensor(\glie) / I$ and~$\Univ(\glie)$ from the universal property of~$\Univ(\glie)$ that the homomorphism~$\Psi'$ factors trough a homorphism of algebras from~$\Tensor(\glie) / I$ to~$A$ if and only if the linear map~$g$ is a homomorphism of Lie~algebras.

	That~$g$ is a homomorphism of Lie~algebras means that
	\[
		g(x) g(y) - g(y) g(x) - g([x,y]) = 0
	\]
	for all~$x, y \in \glie$.
	This is equivalent to the condition
	\[
		\Psi(x) \Psi(y) - \Psi(y) \Psi(x) - \Psi([x,y]_{\glie})
		=
		0
	\]
	for all~$x, y \in \glie$, and furthermore equivalent to the condition
	\[
		\Psi( xy - yx - [x,y]_{\glie} )
		=
		0
	\]
	for all~$x, y \in \glie$.

	We have now seen that an algebra homomorphism~$\Psi$ from~$\Tensor(\glie)$ to some~\algebra{$\kf$}~$A$ factors trough the quotient~$\Tensor(\glie)/I$ if and only if~$\Psi$ annihilates all those elements of~$\Tensor(\glie)$ that are of the form~$xy - yx - [x,y]_{\glie}$ where~$x$ and~$y$ range through~$\glie$.
	This means that the ideal~$I$ needs to be generated by those elements.
	
	We have now altogether seen that the universal enveloping algebra~$\Univ(\glie)$ needs to be constructable as the quotient of the tensor algebra~$\Tensor(\glie)$ by the ideal~$I$ that is generated by all those elements of the form~$x y - y x - [x,y]_{\glie}$ where~$x$ and~$y$ range through~$\glie$.
	We will conversely show in the following \lcnamecref{existence of uea} that this construction will indeed give us the universal enveloping algebra.
\end{fluff}


\begin{proposition}[Existence of the universal enveloping algebra]
	\label{existence of uea}
	Let~$\glie$ be a~\liealgebra{$\kf$}.
	Let~$\Tensor(\glie)$ be the tensor algebra of the underlying vector space of~$\glie$ and let~$I$ the two-sided ideal of~$\Tensor(\glie)$ generated by all elements of the form~$x y - y x - [x,y]_{\glie}$ where~$x$ and~$y$ range through~$\glie$.
	The quotient algebra~$U \defined T(\glie)/I$ together with the~{\linear{$\kf$}} map
	\[
		\iota
		\colon
		\glie
		\to
		\Univ(\glie) \,,
		\quad
		x
		\mapsto
		\class{x}
	\]
	is a universal enveloping algebra for~$\glie$.
\end{proposition}


\begin{proof}
	The map~$\iota$ is~{\linear{$\kf$}}, and it is compatible with the Lie brackets of~$\glie$ and~$\Univ(\glie)$ because
	\[
		[\iota(x), \iota(y)]
		=
		[\class{x}, \class{y}]
		=
		\class{x} \, \class{y} - \class{y} \, \class{x}
		=
		\class{x y - y x}
		=
		\class{[x,y]_{\glie}}
		=
		\iota([x,y]_{\glie})
	\]
	for all~$x, y \in \glie$.
	Given any~\algebra{$\kf$}~$A$ and Lie algebra homomorphism~$\varphi$ from~$\glie$ to~$A$, there exists a unique homorphism of~\algebras{$\kf$}~$\Phi'$ from~$\Tensor(\glie)$ to~$A$ that makes the triangular diagram
	\[
		\begin{tikzcd}
			\glie
			\arrow{r}[above]{\varphi}
			\arrow{d}
			&
			A
			\\
			\Tensor(\glie)
			\arrow[dashed]{ur}[below right]{\Phi'}
			&
			{}
		\end{tikzcd}
	\]
	commute.
	The homomorphism~$\Phi'$ is given by~$\Phi'(x) = \varphi(x)$ for every~$x \in \glie$.
	It follows that
	\begin{align*}
		\Phi'(x y - y x)
		&=
		\Phi'(x) \Phi'(y) - \Phi'(y) \Phi'(x)
		\\
		&=
		\varphi(x) \varphi(y) - \varphi(y) \varphi(x)
		\\
		&=
		[ \varphi(x), \varphi(y) ]
		\\
		&=
		\varphi( [x,y]_{\glie} )
		\\
		&=
		\Phi'( [x,y]_{\glie} )
	\end{align*}
	for all~$x, y \in \glie$.
	The ideal~$I$ is therefore contained in the kernel of the homomorphism~$\Phi'$.
	It follows that there exists a unique homomorphism of algebras~$\Phi$ from~$U$ to~$A$ that makes the diagram
	\[
		\begin{tikzcd}
			\glie
			\arrow{r}[above]{\varphi}
			\arrow{d}
			&
			A
			\\
			\Tensor(\glie)
			\arrow[bend right= 20]{ur}[above left]{\Phi'}
			\arrow{d}[left]{\Pi}
			&
			{}
			\\
			U
			\arrow[dashed, bend right = 30]{uur}[below right]{\Phi}
			&
			{}
		\end{tikzcd}
	\]
	commute, where~$\Pi$ denotes the canonical projection from~$\Tensor(V)$ to~$\Tensor(V)/I$.
	We may add the homomorphism of Lie~algebras~$\iota$ to this diagram.
	We then arrive at the following commutative diagram.
	\[
		\begin{tikzcd}
			\glie
			\arrow{r}[above]{\varphi}
			\arrow{d}
			\arrow[bend right = 55]{dd}[left]{\iota}
			&
			A
			\\
			\Tensor(\glie)
			\arrow[bend right= 20]{ur}[above left]{\Phi'}
			\arrow{d}[left]{\Pi}
			&
			{}
			\\
			U
			\arrow[bend right = 30]{uur}[below right]{\Phi}
			&
			{}
		\end{tikzcd}
	\]
	We have in particular the following commutative subdiagram.
	\[
		\begin{tikzcd}
			\glie
			\arrow{r}[above]{\varphi}
			\arrow{d}[left]{i}
			&
			A
			\\
			U
			\arrow{ur}[below right]{\Phi}
			&
			{}
		\end{tikzcd}
	\]
	We have thus shown that every homomorphism of Lie~algebras~$\varphi$ from~$\glie$ to~$A$ extends along~$\iota$ to a homomorphism of algebra~$\Phi$ from~$U$ to~$A$.
	The algebra~$U$ is generated by the image of~$\iota$ whence the homomorphism of algebras~$\Phi$ is unique with this property.
\end{proof}


\begin{remark}
	The above proof may be summarized as observing that we have bijections
	\begingroup
	\allowdisplaybreaks
	\begin{align*}
		{}&
		\{ \textstyle\text{algebra homomorphisms~$\Phi \colon \Tensor(\glie)/I \to A$} \}
		\\
		\cong{}&
		\{ \text{algebra homomorphisms~$\Phi' \colon \Tensor(\glie) \to A$ with~$I \subseteq \ker(\Phi')$} \}
		\\
		\cong{}&
		\left\{
			\begin{tabular}{c}
				algebra homomorphisms~$\Phi' \colon \Tensor(\glie) \to A$ with  \\
				$\Phi'(x y - y x - [x,y]_{\glie}) = 0$ for all~$x, y \in \glie$
			\end{tabular}
		\right\}
		\\
		\cong{}&
		\left\{
			\begin{tabular}{c}
				algebra homomorphisms~$\Phi' \colon \Tensor(\glie) \to A$ with  \\
				$\Phi'(x) \Phi'(y) - \Phi'(y) \Phi'(x) - \Phi'([x,y]_{\glie}) = 0$ for all~$x, y \in \glie$
			\end{tabular}
		\right\}
		\\
		\cong{}&
		\left\{
			\begin{tabular}{c}
				algebra homomorphisms~$\Phi' \colon \Tensor(\glie) \to A$ with  \\
				$\Phi'(x) \Phi'(y) - \Phi'(y) \Phi'(x) = \Phi'([x,y]_{\glie})$ for all~$x, y \in \glie$
			\end{tabular}
		\right\}
		\\
		\cong{}&
		\left\{
			\begin{tabular}{c}
				{\linear{$\kf$}} maps~$\varphi \colon \glie \to A$ with  \\
				$\varphi(x) \varphi(y) - \varphi(y) \varphi(x) = \varphi([x,y])$ for all~$x, y \in \glie$
			\end{tabular}
		\right\}
		\\
		\cong{}&
		\left\{
			\begin{tabular}{c}
				{\linear{$\kf$}} maps~$\varphi \colon \glie \to A$ with  \\
				$[\varphi(x), \varphi(y)] = \varphi([x,y])$ for all~$x, y \in \glie$ 
			\end{tabular}
		\right\}
		\\
		={}&
		\{ \textstyle\text{Lie~algebra homomorphisms~$\varphi \colon \glie \to A$} \} \,,
	\end{align*}
	\endgroup
	and that these bijections are natural in~$A$.
	This shows that the~\algebra{$\kf$}~$\Tensor(\glie)/I$ represents the correct functor to be the universal enveloping algebra of~$\glie$.
	We can also see that the identity map of~$\Tensor(\glie)/I$ corresponds under the above bijections (for~$A = \Tensor(\glie)/I$) to the specified map~$\iota$ from~$\glie$ to~$\Tensor(\glie)/I$.
\end{remark}



\subsection{Properties}


\begin{convention}
	Let~$\glie$ be a Lie~algebra.
	For every element~$x$ of~$\glie$ the image of~$x$ of~$\Univ(\glie)$ is denoted by~$\class{x}$\glsadd{element in universal enveloping algebra}.
\end{convention}

\subsubsection{Anti-Homomorphisms}

\begin{proposition}
	Let~$\glie$ be a Lie~algebra, let~$\iota$ be the canonical homomorphism of Lie~algebras from~$\glie$ to~$\Univ(\glie)$ and let~$A$ be a~\algebra{$\kf$}.
	We have a well-defined {\onetoonetext} correspondence given by
	\begin{align*}
		\SwapAboveDisplaySkip
		\left\{
			\begin{tabular}{c}
				anti-homomorphisms\index{anti-homomorphism!of Lie algebras} \\
				of Lie~algebras
				$\varphi \colon \glie \to A$
			\end{tabular}
		\right\}
		&\onetoone
		\left\{
			\begin{tabular}{c}
				anti-homomorphisms\index{anti-homomorphism!of algebras} \\
				of algebras
				$\Phi \colon \Univ(\glie) \to A$
			\end{tabular}
		\right\} \,,
		\\
		\Phi \circ \iota
		&\mapsfrom
		\Phi \,.
	\end{align*}
\end{proposition}

\begin{proof}
	We have {\onetoonetext} correspondence given by
	\begin{align*}
		\left\{
			\begin{tabular}{c}
				homomorphisms \\
				of Lie~algebras
				$\varphi \colon \glie \to A^{\op}$
			\end{tabular}
		\right\}
		&\onetoone
		\left\{
			\begin{tabular}{c}
				homomorphisms \\
				of algebras
				$\Phi \colon \Univ(\glie) \to A^{\op}$
			\end{tabular}
		\right\} \,,
		\\
		\Phi \circ \iota
		&\mapsfrom
		\Phi \,.
	\end{align*}
	An anti-homomorphism of Lie~algebras from~$\glie$ to~$A$ is the same as a homomorphim of Lie~algebras from~$\glie$ to~$A^{\op}$, and an anti-homomorphism of algebras from~$\Univ(\glie)$ to~$A$ is the same as a homomorphism of algebras from~$\Univ(\glie)$ to~$A^{\op}$.%
	\footnote{
		We use here implicitely that the opposite Lie algebra of~$A$ is the same as the underlying Lie~algebra of the opposite algebra of~$A$.
		In other words, the operations \enquote{taking the opposite} and \enquote{taking the underlying Lie~algebra} commute.
	}
\end{proof}

\subsubsection{Representations and Modules}

\begin{proposition}
	\label{representations are modules}
	Let~$M$ be a~{\vectorspace{$\kf$}} and let~$\glie$ be a Lie~algebra.
	Let~$\Univ(\glie)$ be the universal enveloping algebra of~$\glie$ and let~$\iota$ be the canonical homorphism of Lie~algebras from~$\glie$ to~$\Univ(\glie)$.
	\begin{enumerate}
		\item
			For every homomorphism of Lie~algebras~$\rho$ from~$\glie$ to~$\gllie(M)$ let~$\widehat{\rho}$ denote the corresponding homomorphism of algebras from~$\Univ(\glie)$ to~$\End_{\kf}(M)$.
			Then the assignments
			\begin{align*}
				\left\{
				\begin{tabular}{c}
					representations\index{representation} \\
					$\rho \colon \glie \to \gllie(M)$
				\end{tabular}
				\right\}
				&\onetoone
				\left\{
				\begin{tabular}{c}
					$\Univ(\glie)$-module structures \\
					$\Rho \colon \Univ(\glie) \to \End_{\kf}(M)$
				\end{tabular}
				\right\}  \,,
				\\
				\rho
				&\mapsto
				\widehat{\rho} \,,
				\\
				\Rho \circ \iota
				&\mapsfrom
				\Rho  \,,
			\end{align*}
			constitute a {\onetoonetext} correspondence.
		\item
			For every anti-homomorphism of Lie~algebras~$\rho$ from~$\glie$ to~$\gllie(M)$ let~$\widehat{\rho}$ denote the corresponding anti-homomorphism of algebras from~$\Univ(\glie)$ to~$\End_{\kf}(M)$.
			Then the assignments
			\begin{align*}
				\left\{
				\begin{tabular}{c}
					right representations \\
					$\rho \colon \glie \to \gllie(M)$
				\end{tabular}
				\right\}
				&\onetoone
				\left\{
				\begin{tabular}{c}
					right $\Univ(\glie)$-module structures \\
					$\Rho \colon \Univ(\glie) \to \End_{\kf}(M)$
				\end{tabular}
				\right\}  \,,
				\\
				\rho
				&\mapsto
				\widehat{\rho} \,,
				\\
				\Rho \circ \iota
				&\mapsfrom
				\Rho  \,,
			\end{align*}
			constitute a {\onetoonetext} correspondence.
		\qed
	\end{enumerate}
\end{proposition}

\begin{fluff}
	Let~$\glie$ be a Lie~algebra.
	If~$M$ is a left~\module{$\Univ(\glie)$}, then the corresponding action of~$\glie$ on~$M$ is given by
	\[
		x \act m
		=
		\class{x} \cdot m
	\]
	for all~$x \in \glie$ and~$m \in M$.
	Similarly, if~$M$ is a right~\module{$\Univ(\glie)$}, then the corresponding right action of~$\glie$ on~$M$ is given by
	\[
		m \act x
		=
		m \cdot \class{x}
	\]
	for all~$x \in \glie$ and~$m \in M$.
\end{fluff}

\begin{remark}
	Let~$\glie$ be a Lie~algebra and let~$\Univ(\glie)$ be the universal enveloping algebra of~$\glie$.
	\Cref{representations are modules} shows that representations of~$\glie$ are the same as~{\modules{$\Univ(\glie)$}}.
	We get from this correspondence an isomorphism of categories between~$\cRep{\glie}$ and~$\cMod{\Univ(\glie)}$.
\end{remark}

\subsubsection{Functoriality}

\begin{proposition}[Functoriality of the universal enveloping algebra]
	\label{functoriality of universal enveloping algebra}
	Let~$\glie$,~$\hlie$ and~$\klie$ be Lie~algebras.
	\begin{enumerate}
		\item
			For every homomorphism of Lie~algebras~$\varphi$ from~$\glie$ to~$\hlie$ there exists a unique homomorphism of algebras~$\Univ(\varphi)$\glsadd{universal enveloping algebra on morphisms}\index{induced homomorphism!of universal enveloping algebras} from~$\Univ(\glie)$ to~$\Univ(\hlie)$ that makes the following square diagram commute.
			\[
				\begin{tikzcd}[column sep = large]
					\glie
					\arrow{r}[above]{\varphi}
					\arrow{d}[left]{\iota_{\glie}}
					&
					\hlie
					\arrow{d}[right]{\iota_{\hlie}}
					\\
					\Univ(\glie)
					\arrow[dashed]{r}[below]{\Univ(\varphi)}
					&
					\Univ(\hlie)
				\end{tikzcd}
			\]
		\item
			It holds that~$\Univ(\id_{\glie}) = \id_{\Univ(\glie)}$.
		\item
			It holds for overy homomorphisms of Lie~algebras~$\varphi$ from~$\glie$ to~$\hlie$ and every homomorphism of Lie~algebra~$\psi$ from~$\hlie$ to~$\klie$ that
			\[
				\Univ( \psi \circ \varphi )
				=
				\Univ( \psi ) \circ \Univ( \varphi ) \,.
			\]
	\end{enumerate}
\end{proposition}


\begin{proof}
	\leavevmode
	\begin{enumerate}
		\item
			The composite~$\iota_{\hlie} \circ \varphi$ is a homomorphism of Lie~algebras from~$\glie$ to~$\Univ(\hlie)$.
			By the universal property of the universal enveloping algebra~$\Univ(\glie)$ there exists a unique homomorphism of algebras~$\Univ(\varphi)$ from~$\Univ(\glie)$ to~$\Univ(\hlie)$ with~$\Univ(\varphi) \circ \iota_{\glie} = \iota_{\hlie} \circ \varphi$.
		\item
			The square diagram
			\[
				\begin{tikzcd}[column sep = huge]
					\glie
					\arrow{r}[above]{\id_{\glie}}
					\arrow{d}
					&
					\glie
					\arrow{d}
					\\
					\Univ(\glie)
					\arrow[dashed]{r}[below]{\id_{\Univ(\glie)}}
					&
					\Univ(\glie)
				\end{tikzcd}
			\]
			commutes, which shows that the identity homomorphism~$\id_{\Univ(\glie)}$ satisfies the defining property of the induced algebra homomorphism~$\Univ( \id_{\glie} )$.
		\item
			We have the following commutative diagram:
			\[
				\begin{tikzcd}[column sep = large]
					\glie
					\arrow[dashed, bend left = 40]{rr}[above]{\psi \circ \varphi}
					\arrow{r}[above]{\varphi}
					\arrow{d}
					&
					\hlie
					\arrow{r}[above]{\psi}
					\arrow{d}
					&
					\klie
					\arrow{d}
					\\
					\Univ(\glie)
					\arrow{r}[below]{\Univ(\varphi)}
					\arrow[dashed, bend right = 40]{rr}[below]{\Univ(\psi) \circ \Univ(\varphi)}
					&
					\Univ(\hlie)
					\arrow{r}[below]{\Univ(\psi)}
					&
					\Univ(\klie)
				\end{tikzcd}
			\]
			The commutativity of the outer square diagram
			\[
				\begin{tikzcd}[column sep = huge]
					\glie
					\arrow{r}[above]{\psi \circ \varphi}
					\arrow{d}
					&
					\klie
					\arrow{d}
					\\
					\Univ(\glie)
					\arrow[dashed]{r}[below]{\Univ(\psi) \circ \Univ(\varphi)}
					&
					\Univ(\klie)
				\end{tikzcd}
			\]
			shows that the composite~$\Univ(\psi) \circ \Univ(\varphi)$ satisfies the defining property of the induced algebra homomorphism~$\Univ(\psi \circ \varphi)$.
		\qedhere
	\end{enumerate}
\end{proof}


\begin{remark}
	\Cref{functoriality of universal enveloping algebra} shows that the assignment~$\glie \mapsto \Univ(\glie)$ of a Lie~algebra~$\glie$ to its universal eveloping algebra~$\Univ(\glie)$ can be extended to a (covariant) functor~$\Univ$ from~$\cLie{\kf}$ to~$\cAlg{\kf}$.
	The universal property of the universal enveloping algebra states that the functor~$\Univ$ is left adjoint\index{adjunction} to the forgetful functor from~$\cAlg{\kf}$ to~$\cLie{\kf}$, which assigns to each~{\algebra{$\kf$}} its underlying Lie~algebra.
\end{remark}


\begin{remark}
	\label{actions on the universal enveloping algebra}
	Let~$\glie$ be a Lie~algebra.
	Its universal enveloping algebra~$\Univ(\glie)$ can be made into a representation of~$\glie$ in two ways.
	\begin{enumerate}
		\item
			The algebra~$\Univ(\glie)$ becomes a left module over itself via left multiplication.
			This~\module{$\Univ(\glie)$} structure corresponds to an action of~$\glie$ on~$\Univ(\glie)$ given by
			\[
				x \act t
				=
				\class{x} \cdot t
			\]
			for all~$x \in \glie$ and~$t \in \Univ(\glie)$.
		\item
			For the second action let~$\iota$ denote the canonical homomorphism from~$\glie$ to~$\Univ(\glie)$.
			The adjoint representation of~$\Univ(\glie)$ is a homomorphism of Lie~algebras from~$\Univ(\glie)$ to~$\gllie(\Univ(\glie))$.
			We can pull-back this representation via~$\iota$ to an action of~$\glie$ on~$\Univ(\glie)$, given by
			\[
				x \act t
				=
				[ \class{x}, t ]
			\]
			for all~$x \in \glie$ and~$t \in \Univ(\glie)$.
			Unter this section action, the homomorphism of Lie~algebras~$\iota$ becomes a homomorphism of representations from the adjoint representation of~$\glie$ to~$\Univ(\glie)$.
	\end{enumerate}
\end{remark}


\begin{definition}
	Let~$\glie$ be a Lie~algebra.
	The first action from \cref{actions on the universal enveloping algebra} is the \defemph{regular action} of~$\glie$ on~$\Univ(\glie)$, and the second action is the \defemph{adjoint action} of~$\glie$ on~$\Univ(\glie)$.
\end{definition}

\subsubsection{Functoriality for Anti-Homomorphisms}

\begin{proposition}
	\label{functoriality of universal enveloping algebra for anti-homomorphisms}
	Let~$\glie$,~$\hlie$, and~$\klie$ be Lie~algebras.
	\begin{enumerate}
		\item
			Let~$\varphi$ be an anti-homomorphism of Lie~algebras from~$\glie$ to~$\hlie$.
			There exists a unique anti-homomorphism of algebras~$\widetilde{\Univ}(\varphi)$\index{induced anti-homomorphism of universal enveloping algebras} from~$\glie$ to~$\hlie$ that makes the following square diagram commute.
			\[
				\begin{tikzcd}[column sep = large]
					\glie
					\arrow{r}[above]{\varphi}
					\arrow{d}[left]{\iota_{\glie}}
					&
					\hlie
					\arrow{d}[right]{\iota_{\hlie}}
					\\
					\Univ(\glie)
					\arrow[dashed]{r}[above]{\widetilde{\Univ}(\varphi)}
					&
					\Univ(\hlie)
				\end{tikzcd}
			\]
		\item
			Let~$\varphi$ be a map from~$\glie$ to~$\hlie$ and let~$\psi$ be a map from~$\hlie$ to~$\klie$.
			\begin{enumerate}
				\item
					If~$\varphi$ and~$\psi$ are homomorphisms of Lie~algebras, then their composite~$\psi \circ \varphi$ is again a homomorphism of Lie~algebras and~$\Univ(\psi \circ \varphi) = \Univ(\psi) \circ \Univ(\varphi)$.
				\item
					If~$\varphi$ is a homomorphisms of Lie~algebras and~$\psi$ is an anti-homomorphism of Lie~algebras, then their composite~$\psi \circ \varphi$ is an anti-homomorphism and~$\widetilde{\Univ}(\psi \circ \varphi) = \widetilde{\Univ}(\psi) \circ \Univ(\varphi)$.
				\item
					If~$\varphi$ is an anti-homomorphisms of Lie~algebras and~$\psi$ is a homomorphism of Lie~algebras, then their composite~$\psi \circ \varphi$ is an anti-homomorphism and~$\widetilde{\Univ}(\psi \circ \varphi) = \Univ(\psi) \circ \widetilde{\Univ}(\varphi)$.
				\item
				If~$\varphi$ and~$\psi$ are anti-homomorphisms of Lie~algebras, then their composite~$\psi \circ \varphi$ is a homomorphism of Lie~algebras and~$\Univ(\psi \circ \varphi) = \widetilde{\Univ}(\psi) \circ \widetilde{\Univ}(\varphi)$.
			\end{enumerate}
		\item
			\label{induced homomorphism same for homomorphisms and anti-homomorphisms}
			Let~$\varphi$ be a map from~$\glie$ to~$\hlie$ that is both a homomorphism of Lie~algebras and an anti-homomorphism of Lie~algebras.
			Then the induced anti-homomorphism of algebras~$\Univ(\varphi)$ agrees with the induced anti-homomorphism of algebras~$\widetilde{\Univ}(\varphi)$.
	\end{enumerate}
\end{proposition}


\begin{proof}
	\leavevmode
	\begin{enumerate}
		\item
			The composite~$\iota_{\hlie} \circ \varphi$ is an anti-homomorphism of Lie~algebras from~$\glie$ to~$\Univ(\hlie)$, and thus a homomorphism of Lie~algebras from~$\glie$ to~$\Univ(\hlie)^{\op}$.
			It thus extends uniquely to a homomorphism of algebras~$\Univ(\varphi)$ from~$\Univ(\glie)$ to~$\Univ(\hlie)^{\op}$ such that~$\Univ(\varphi) \circ \iota_{\glie} = \iota_{\hlie} \circ \varphi$.
			We can now regard~$\Univ(\varphi)$ as an anti-homomorphism of algebras~$\widetilde{U}(\varphi)$ from~$\Univ(\glie)$ to~$\Univ(\hlie)$.
		\item
			If~$\varphi$ and~$\psi$ are homomorphism of Lie~algebras, then this is known from \cref{functoriality of universal enveloping algebra}.
			The other cases can be shown in the same way is in the proof of \cref{functoriality of universal enveloping algebra}.
		\item
			Suppose first that the Lie~algebra~$\hlie$ is abelian.
			Its universal enveloping algebra~$\Univ(\hlie)$ is commutative because it is generated by the image of~$\hlie$ in~$\Univ(\hlie)$.
			The anti-homomorphism of algebras~$\widetilde{\Univ}(\varphi)$ is therefore a homomorphism of algebras, and it satisfies the defining property of the induced homomorphism of algebras~$\Univ(\varphi)$.
			We thus find that~$\widetilde{\Univ}(\varphi)$ equals~$\Univ(\varphi)$.

			For the general case let~$\ilie$ be the image of~$\varphi$.
			This is a Lie~subalgebra of~$\hlie$ because~$\varphi$ is a homomorphism of Lie~algebras.
			Let~$\varphi'$ be the corestriction of~$\varphi$ to a map from~$\glie$ to~$\varphi'$ and let~$\iota$ be the inclusion map from~$\ilie$ to~$\hlie$.
			The map~$\varphi'$ is again both a homomorphism of Lie~algebras and an anti-homomorphism of Lie~algebras, and the inclusion map~$\iota'$ is a homomorphism of Lie~algebras.
			The Lie~algebra~$\ilie$ is abelian because
			\[
				[\varphi(x), \varphi(y)]
				=
				\varphi( [x,y] )
				=
				[\varphi(y), \varphi(x)]
			\]
			for all~$x, y \in \glie$.
			It follows from the previously considered special case that~$\Univ(\varphi') = \widetilde{\Univ}(\varphi')$.
			It finally follows that
			\[
				\widetilde{\Univ}(\varphi)
				=
				\widetilde{\Univ}(\iota \circ \varphi')
				=
				\Univ(\iota) \circ \widetilde{\Univ}(\varphi')
				=
				\Univ(\iota) \circ \Univ(\varphi')
				=
				\Univ(\iota \circ \varphi')
				=
				\Univ(\varphi) \,,
			\]
			as desired.
		\qedhere
	\end{enumerate}
\end{proof}


\begin{notation}
	Let~$\glie$ and~$\hlie$ be two Lie~algebras and let~$\varphi$ be a anti-homomorphism of Lie~algebras from~$\glie$ to~$\hlie$.
	Instead of~$\widetilde{\Univ}(\varphi)$ we will write~$\Univ(\varphi)$\glsadd{universal enveloping algebra on morphisms}.
	This does not need to conflicts in the case that~$\varphi$ is also a homomorphism of Lie~algebras, as shown in part~\ref{induced homomorphism same for homomorphisms and anti-homomorphisms} of \cref{functoriality of universal enveloping algebra for anti-homomorphisms}.
\end{notation}


\begin{corollary}
	\label{anti-isomorphism of lie-algebras induces anti-isomorphism of algebras}
	Let~$\glie$ and~$\hlie$ be two Lie~algebras and let~$\varphi$ be an anti-isomorphism of Lie~algebras from~$\glie$ to~$\hlie$.
	The induced anti-homorphism of algebras~$\Univ(\varphi)$ from~$\Univ(\glie)$ to~$\Univ(\hlie)$ is an isomorphism of algebras.
	\qed
\end{corollary}


\begin{example}
	\label{construction of antipode}
	Let~$\glie$ be a Lie~algebra.
	The map
	\[
		\sigma
		\colon
		\glie
		\to
		\glie \,,
		\quad
		x
		\mapsto
		-x
	\]
	is an anti-involution of Lie~algebras of~$\glie$.
	The induced anti-homomorphism of Lie~algebras~$\Univ(\sigma)$ is an anti-involution of algebras of~$\Univ(\glie)$.
\end{example}


\begin{definition}
	Let~$\glie$ be a Lie~algebra.
	In the situation of \cref{construction of antipode}, the induced anti-involution~$\Univ(\sigma)$ of~$\Univ(\glie)$ is the \defemph{antipode}\index{antipode} of~$\Univ(\glie)$.
	It is denoted by~$S_{\glie}$, or simply by~$S$\glsadd{antipode}.
\end{definition}


\subsubsection{Derivations}

\begin{proposition}
	\label{extending derivation to universal enveloping algebra}
	Let~$\glie$ be a Lie~algebra.
	Every derivation\index{derivation!of Lie algebra} of~$\glie$ extends uniquely to a derivation of~$\Univ(\glie)$.
	More explicitely, there exists for every Lie~algebra derivation~$\delta$ of~$\glie$ a unique algebra derivation~$\Delta$ of~$\Univ(\glie)$ such that the following square diagram commutes.
	\[
		\begin{tikzcd}
			\glie
			\arrow{r}[above]{\delta}
			\arrow{d}
			&
			\glie
			\arrow{d}
			\\
			\Univ(\glie)
			\arrow[dashed]{r}[above]{\Delta}
			&
			\Univ(\glie) \,.
		\end{tikzcd}
	\]
\end{proposition}


\begin{fluff}
	To prove \cref{extending derivation to universal enveloping algebra} we need some preparation.
	To employ the universal property of the universal enveloping algebra we want to be able to translate between derivations and homomorphisms.
	This is what the next \lcnamecref{translating between derivations and homomorphisms} is about.
\end{fluff}


\begin{lemma}
	\label{translating between derivations and homomorphisms}
	\leavevmode
	\begin{enumerate}
		\item
			Let~$A$ be a~\algebra{$\kf$} and let~$B$ be the~\algebra{$\kf$}
			\[
				B
				\defined
				\begin{pmatrix}
					A & A \\
					0 & A
				\end{pmatrix} \,.
			\]
			A map~$\Delta$ from~$A$ to~$A$ is a derivation of~$A$ if and only if the map
			\[
				\Phi
				\colon
				A
				\to
				B \,,
				\quad
				a
				\mapsto
				\begin{pmatrix}
					a & \Delta(a) \\
					0 & a
				\end{pmatrix}
			\]
			is a homomorphism of algebras.
		\item
			Let~$\glie$ be a Lie~algebra and let
			\[
				B
				\defined
				\begin{pmatrix}
					\Univ(\glie)  & \Univ(\glie)  \\
					0             & \Univ(\glie)
				\end{pmatrix} \,.
			\]
			We regard~$\Univ(\glie)$ as a representation of~$\glie$ via the adjoint action of~$\glie$ on~$\Univ(\glie)$.
			A map~$\delta$ from~$\glie$ to~$\Univ(\glie)$ is a derivation if and only if the map
			\[
				\varphi
				\colon
				\glie
				\to
				B \,,
				\quad
				x
				\mapsto
				\begin{pmatrix}
					\class{x} & \delta(x) \\
					0         & \class{x}
				\end{pmatrix}
			\]
			is a homomorphism of Lie~algebras
	\end{enumerate}
\end{lemma}


\begin{proof}
	\leavevmode
	\begin{enumerate}
		\item
			The map~$\Phi$ is linear if and only if the map~$\Delta$ is linear.
			We have
			\[
				\Phi(a) \cdot \Phi(b)
				=
				\begin{pmatrix}
					a & \Delta(a) \\
					0 & a
				\end{pmatrix}
				\begin{pmatrix}
					b & \Delta(b) \\
					0 & b
				\end{pmatrix}
				=
				\begin{pmatrix}
					ab  & \Delta(a) b + a \Delta(b) \\
					0   & ab
				\end{pmatrix}
			\]
			for all~$a, b \in A$.
			It follows that the map~$\Phi$ is multiplicaitve if and only if~$\Delta(ab) = \Delta(a) b + a \Delta(b)$ for all~$a, b \in A$.
			If~$\Delta$ is a derivation then we have~$\Phi(1) = 1$.
			This shows altogether that~$\Delta$ is a derivation if and only if~$\Phi$ is a homomorphism of algebras.
		\item
			We have
			\begin{align*}
				[ \varphi(x), \varphi(y) ]
				&=
				\Biggl[
					\begin{pmatrix}
						\class{x} & \delta(x) \\
						0         & \class{x}
					\end{pmatrix},
					\begin{pmatrix}
						\class{y} & \delta(y) \\
						0         & \class{y}
					\end{pmatrix}
				\Biggr]
				\\
				&=
				\begin{pmatrix}
					\class{x} & \delta(x) \\
					0         & \class{x}
				\end{pmatrix}
				\begin{pmatrix}
					\class{y} & \delta(y) \\
					0         & \class{y}
				\end{pmatrix}
				-
				\begin{pmatrix}
					\class{y} & \delta(y) \\
					0         & \class{y}
				\end{pmatrix}
				\begin{pmatrix}
					\class{x} & \delta(x) \\
					0         & \class{x}
				\end{pmatrix}
				\\
				&=
				\begin{pmatrix}
					\class{xy}  & \class{x} \delta(y) + \delta(x) \class{y} \\
					0           & \class{xy}
				\end{pmatrix}
				-
				\begin{pmatrix}
					\class{yx}  & \class{y} \delta(x) + \delta(y) \class{x} \\
					0           & \class{yx}
				\end{pmatrix}
				\\
				&=
				\begin{pmatrix}
					\class{x}\class{y} - \class{y}\class{x} & \class{x} \delta(y) - \delta(y) \class{x} + \delta(x) \class{y} - \class{y} \delta(x) \\
					0                                       & \class{x}\class{y} - \class{y}\class{x}
				\end{pmatrix}
				\\
				&=
				\begin{pmatrix}
					\class{[x,y]} & [\class{x}, \delta(y)] - [\class{y}, \delta(x)] \\
					0             & \class{[x,y]}
				\end{pmatrix}
				\\
				&=
				\begin{pmatrix}
					\class{[x,y]} & x \act \delta(y) -  y \act \delta(x) \\
					0             & \class{[x,y]}
				\end{pmatrix}
			\end{align*}
			for all~$x, y \in \glie$.
			It follows that the map~$\varphi$ is a homomorphism of Lie~algebras if and only if~$\delta([x,y]) = x \act \delta(x) - y \act \delta(x)$ for all~$x, y \in \glie$, i.e. if and only if the map~$\delta$.
		\qedhere
	\end{enumerate}
\end{proof}


\begin{proof}[Proof of \cref{extending derivation to universal enveloping algebra}]
	The uniqueness of~$\Delta$ follows from \cref{dervation is uniquely determined by algebra generators} because the universal enveloping algebra~$\Univ(\glie)$ is generated by the image of~$\glie$ in~$\Univ(\glie)$.

	Let~$B$ be the~\algebra{$\kf$} given by
	\[
		B
		\defined
		\begin{pmatrix}
			\Univ(\glie)  & \Univ(\glie) \\
			0             & \Univ(\glie)
		\end{pmatrix} \,.
	\]
	We regard~$\Univ(\glie)$ as a representation of~$\glie$ via the adjoint action of~$\glie$ on~$\Univ(\glie)$.
	The canonical homomorphism of Lie~algebras from~$\glie$ to~$\Univ(\glie)$ is a homomorphism of representations, whence it follows that the map
	\[
		\glie
		\to
		\Univ(\glie) \,,
		\quad
		x
		\mapsto
		\class{\delta(x)}
	\]
	is again a derivation.
	It follows from \cref{translating between derivations and homomorphisms} that the map
	\[
		\varphi
		\colon
		\glie
		\to
		B \,,
		\quad
		x
		\mapsto
		\begin{pmatrix}
			\class{x} & \class{\delta(x)} \\
			0         & \class{x}
		\end{pmatrix}
	\]
	is a homomorphism of Lie~algebras because~$\delta$ is a derivation of~$\glie$.
	It follows from the universal property of the universal enveloping algebra~$\Univ(\glie)$ that the homomorphism of Lie~algebras~$\varphi$ extends uniquely to a homomorphism of algebras~$\Phi$ from~$\Univ(\glie)$ to~$B$.
	This homomorphism~$\Phi$ is of the form
	\[
		\Phi(y)
		=
		\begin{pmatrix}
			\Phi_1(y) & \Delta(y) \\
			0         & \Phi_2(y)
		\end{pmatrix}
		\qquad
		\text{for every~$y \in \Univ(\glie)$}
	\]
	for some unique linear maps
	\[
		\Phi_1, \Phi_2, \Delta
		\colon
		\Univ(\glie)
		\to
		\Univ(\glie) \,.
	\]
	The maps~$\Phi_1$ and~$\Phi_2$ are homomorphisms of algebras because~$\Phi$ is a homomorphism of algebras.
	They satisfy the equalities~$\Phi_1(\class{x}) = \varphi(x) = \class{x}$ and~$\Phi_2(\class{x}) = \varphi(x) = \class{x}$ for every~$x \in \glie$.
	It follows that~$\Phi_1(y) = y$ and~~$\Phi_2(y) = y$ for every~$y \in \Univ(\glie)$ because~$\glie$ generates~$\Univ(\glie)$ as an algebra.
	The homomorphism~$\Phi$ is thus of the form
	\[
		\Phi(y)
		=
		\begin{pmatrix}
			y & \Delta(y) \\
			0 & y
		\end{pmatrix}
		\qquad
		\text{for every~$y \in \Univ(\glie)$.}
	\]
	It follows from \cref{translating between derivations and homomorphisms} that the linear map~$\Delta$ is an algebra derivation of~$\Univ(\glie)$.
	This darivation satisfies the equality~$\Delta(\class{x}) = \class{\delta(x)}$ for every~$x \in \glie$ because~$\Phi$ is an extension of~$\varphi$.
	In other words,~$\Delta$ is an extension of~$\delta$.
\end{proof}


