\section{Nilpotent and Solvable Lie~Algebras}





\subsection{Definition, Examples and Properties}


\begin{definition}
  Let~$A$ be an associative~{\algebra{$\kf$}}.
  An element~$a \in A$ is \defemph{nilpotent}\index{nilpotent} if~$a^n = 0$ for some~$n \geq 1$.
  Given a Lie~algebra~$\glie$ an element~$x \in \glie$ is called \defemph{\adnilpotent} if the endomorphism~$\ad(x) \in \End_{\kf}(\glie)$ is nilpotent.
\end{definition}


\begin{lemma}
  \label{nilpotent implies ad-nilpotent}
  If~$A$ is an associative~{\algebra{$\kf$}} and~$x \in A$ is nilpotent then~$x$ is also {\adnilpotent}.
\end{lemma}

\begin{proof}
  Let~$\lambda_x \colon A \to A$,~$a \mapsto xa$ denote the left multiplication by~$x$ and~$\rho_x \colon A \to A$,~$a \mapsto ax$ the right multiplication by~$x$.
  Both~$\lambda_x$ and~$\rho_x$ are nilpotent because~$x$ is nilpotent.
  The endomorphisms~$\lambda_x$ and~$\rho_x$ commute because~$A$ is associative.
  Hence~$\ad(x) = \lambda_x - \rho_x$ is the sum (or rather difference) of two commuting nilpotent endomorphisms, and therefore again nilpotent.
\end{proof}


\begin{definition}
  Let~$\glie$ be a Lie~algebra.
  \begin{enumerate}
    \item
      Let~$\glie^0 \defined \glie$ and inductively~$\glie^{i+1} \defined [\glie, \gls*{central series}]$ for every~$i \geq 0$.
      The decreasing sequence
      \[
        \glie
        =
        \glie^0
        \supseteq
        \glie^1
        \supseteq
        \glie^2
        \supseteq
        \glie^3
        \supseteq
        \dotsb
      \]
      is the \defemph{central series}\index{central series}\index{series!central} of~$\glie$.
      The Lie~algebra~$\glie$ is \defemph{nilpotent}\index{nilpotent!Lie algebra}\index{Lie algebra!nilpotent} if~$\glie^n = 0$ for~$n$ sufficiently large.
    \item
      Let~$\glie^{(0)} \defined \glie$ and~$\glie^{(i+1)} \defined [\glie^{(i)}, \gls*{derived series}]$ for every~$i \geq 0$.
      The decreasing sequence
      \[
        \glie
        =
        \glie^{(0)}
        \supseteq
        \glie^{(1)}
        \supseteq
        \glie^{(2)}
        \supseteq
        \glie^{(3)}
        \supseteq
        \dotsb
      \]
      is the \defemph{derived series}\index{derived series}\index{series!derived} of~$\glie$.
      The Lie~algebra~$\glie$ is \defemph{solvable}\index{solvable}\index{Lie algebra!solvable} if~$\glie^{(n)} = 0$ for~$n$ sufficiently large.
  \end{enumerate}
\end{definition}


\begin{remark}
  Let~$\glie$ be a Lie~algebra.
  \begin{enumerate}
    \item
      Both the central series~$(\glie^i)_{i \geq 0}$ and the derived series~$(\glie^{(i)})_{i \geq 0}$ consist of ideals of~$\glie$.
    \item
      If~$\glie$ is nilpotent then it is also solvable because~$\glie^{(i)} \subseteq \glie^i$ for every~$i \geq 0$.
    \item
      Our indexing convention is choosen so that~$(\glie^{(i)})^{(j)} = \glie^{(i+j)}$ for all~$i, j \geq 0$.
    \item
      It follows that~$\glie$ is solvable if and only if~$\glie^{(i)}$ is solvable for some~$i \geq 0$, if and only if~$\glie^{(i)}$ is solvable for every~$i \geq 0$.
%     \item
%       One can define more generally for every ideal~$I$ in~$\glie$ the central series series~$(I^i)_{i \geq 0}$ of~$I$ in~$\glie$ by~$I^0 \defined \glie$ and~$I^{i+1} \defined [\glie, I^i]$ for every~$i \geq 0$.
%       Then
%       \[
%         \glie
%         =
%         I^0
%         \supseteq
%         I^1
%         \supseteq
%         I^2
%         \supseteq
%         I^3
%         \supseteq
%         \dotsb
%       \]
%       is a decreasing sequence of ideals in~$\glie$.
%       Note that the ideal~$I$ itself does not have to appear in this sequence;
%       the term~$I^1$ is contained in~$I$ (as are all following terms) but does not have to coincide with~$I$.
%       (Suppose for example that~$\glie$ is nonabelian and that~$I$ is nonzero but contained in the center~$\centerlie(\glie)$.
%       Then~$I \neq \glie = I^0$ but already~$I^1 = [I, \glie] = 0$.)
  \end{enumerate}
\end{remark}


\begin{examples}
  \label{examples for solvable and nilpotent}
  \leavevmode
  \begin{enumerate}
    \item
      The Lie~algebra of upper triangular matrices~$\tlie_n(\kf)$ is solvable but not nilpotent.
    \item
      The Lie~algebra of strictly upper triangular matrices~$\nlie_n(\kf)$ is nilpotent, and hence also solvable.
    \item
      If~$n \geq 2$ then the special linear Lie~algebra~$\sllie_n(\Complex)$ is simple, and hence~$[\sllie_n(\Complex), \sllie_n(\Complex)] = \sllie_n(\Complex)$.
      Thus~$\sllie_n(\Complex)$ is neither solvable nor nilpotent.
    \item
      The general linear Lie~algebra~$\gllie_n(\Complex)$ is not solvable because
      \[
        \gllie_n(\Complex)^{(1)}
        =
        [\gllie_n(\Complex), \gllie_n(\Complex)]
        =
        \sllie_n(\Complex)
      \]
      is not solvable.
    \item
      If~$\glie$ is abelian then~$\glie$ is nilpotent (with~$\glie^1 = 0$) and therefore also solvable.
    \item
      Every {\onedimensional} Lie~algebra is abelian, thus nilpotent and hence solvable.
      The same goes for the {\twodimensional} abelian Lie~algebra.
      The {\twodimensional} non-abelian Lie~algebra~$\glie$ admits a basis~$x$,~$y$ with~$[x,y] = x$.
      Then~$\glie^{(i)} = 0$ for every~$i \geq 2$ but~$\glie^i = \kf x$ for every~$i \geq 1$.
      The Lie~algebra~$\glie$ is thus solvable but not nilpotent.
    \item
      The \defemph{Heisenberg Lie~algebra}\index{Heisenberg Lie algebra}\index{Lie algebra!Heisenberg} consists of a real vector space with basis~$P_1, \dotsc, P_n, Q_1, \dotsc, Q_n, C$ together with the Lie~bracket satisfying the relations
      \begin{itemize}
        \item
          $[P_i, P_j] = 0$ and~$[Q_i, Q_j] = 0$ for all~$i, j = 1, \dotsc, n$,
        \item
          $[P_i, C] = [Q_i, C] = 0$ for all~$i = 1, \dotsc, n$,
        \item
          $[P_i, Q_j] = \delta_{ij} C$ for all~$i,j = 1, \dotsc, n$.
      \end{itemize}
      This defines a nilpotent Lie~algebra.
      
      As a faithful representation of the Heisenberg Lie~algebra (and thus a realization of the Heisenberg Lie~algebra via endomorphisms) we can consider the polynomial ring~$V = \Real[x_1, \dotsc, x_n]$.
      Let~$p_1, \dotsc, p_n, q_1, \dotsc, q_n, c \in \gllie(V)$ so that~$p_i = \partial_i$ is the~{\howmanyth{$i$}} partial derivate,~$q_i = \partial_i$ is the multiplication with~$x_i$ and~$c = \id_V$.
      These endomorphisms are linearly independent and satisfy the desired relations.
      It follows that~$\hlie = \vspan_\kf(p_1, \dotsc, p_n, q_1, \dotsc, q_n, c)$ is a Lie subalgebra of~$\gllie(V)$ that is isomorphic to the Heisenberg Lie~algebra.
  \end{enumerate}
\end{examples}


\begin{proposition}[Properites of solvability]
  \label{properties of solvable and nilpotent}
  Let~$\glie$ be a Lie~algebra.
  \begin{enumerate}
    \item
      If~$\hlie$ is another Lie~algebra and~$f \colon \glie \to \hlie$ is a homomorphism of Lie~algebras then
      \[
        f(\glie)^{(i)}
        =
        f(\glie^{(i)})
      \]
      for every~$i \geq 0$.
    \item
      If the Lie~algebra~$\glie$ is solvable then any Lie~subalgebra~$\hlie$ of~$\glie$ is again solvable.
    \item
      If~$\glie$ is solvable and~$I$ is an ideal in~$\glie$ then the quotient Lie~algebra~$\glie/I$ is again solvable.
    \item
      If~$I$ is an ideal in~$\glie$ such that both~$I$ and~$\glie/I$ is solvable then~$\glie$ is solvable.
  \end{enumerate}
\end{proposition}


\begin{proof}
  \leavevmode
  \begin{enumerate}
    \item
      This can be shown by induction because~$f([X,Y]) = [f(X), f(Y)]$ for any two subsets~$X, Y \subseteq \glie$.
    \item
      The inclusion~$\hlie \inclusion \glie$ is a Lie algebra homomorphism, hence~$\hlie^{(i)} \subseteq \glie^{(i)}$ for every~$i \geq 0$.
      It thus follows from~$\glie^{(n)} = 0$ for~$n$ sufficiently large that also~$\hlie^{(n)} = 0$.
    \item
      The canonical projection~$\pi \colon \glie \to \glie/I$ is a Lie algebra homomorphism, and hence
      \[
        (\glie/I)^{(i)}
        =
        \pi(\glie)^{(i)}
        =
        \pi(\glie^{(i)})
      \]
      for every~$i \geq 0$.
      It thus follows from~$\glie^{(n)} = 0$ for~$n$ sufficietly large that also~$(\glie/I)^{(n)} = 0$.
    \item
      It follows from the solvability of~$\glie/I$ that~$(\glie/I)^{(n)} = 0$ for~$n$ sufficiently large.
      This means that
      \[
        0
        =
        (\glie/I)^{(n)}
        =
        \pi(\glie)^{(n)}
        =
        \pi(\glie^{(n)})  \,,
      \]
      so that~$\glie^{(n)} \subseteq I$.
      It follows from the solvability of~$I$ that~$I^{(m)} = 0$ for~$m$ sufficiently large.
      Then
      \[
        \glie^{(n+m)}
        =
        (\glie^{(n)})^{(m)}
        \subseteq
        I^{(m)}
        =
        0
      \]
      and hence~$\glie^{(n+m)} = 0$.
    \qedhere
  \end{enumerate}
\end{proof}


\begin{corollary}
  \label{solvable via ses}
  If~$I$ is an ideal in a Lie~algebra~$\glie$ then~$\glie$ is solvable if and only if both~$I$ and~$\glie/I$ are solvable
  Hence in a short exact sequence
  \[
    0
    \to
    I
    \to
    \glie
    \to
    \glie/I
    \to
    0
  \]
  the middle term is solvable if and only if the two outer terms are solvable.
  \qed
\end{corollary}


\begin{corollary}
  \label{sum of solvable ideals is solvable}
  Let~$\glie$ be a Lie~algebra and let~$I$ and~$J$ be two solvable ideals in~$\glie$.
  Then their sum~$I + J$ is again solvable.
\end{corollary}


\begin{proof}
  We consider the short exact sequence
  \[
    0
    \to
    I
    \to
    I+J
    \to
    (I+J)/I
    \to
    0 \,.
  \]
  The terms~$I$ and~$(I+J)/I \cong J/(I \cap J)$ are both solvable (the second one because~$J$ is solvable) and hence the middle term~$I+J$ is solvable.
\end{proof}


\begin{definition}
 Let~$\glie$ be a finite dimensional Lie~algebra.
 It follows from \cref{sum of solvable ideals is solvable} that~$\glie$ contains a unique maximal solvable ideal. This maximal solvable ideal is the \defemph{radical}\index{radical!of a Lie algebra} of~$\glie$ and is denoted by~\gls*{radical}.
\end{definition}


\begin{proposition}[Properties of nilpotence]
  \label{properties of nilpotence}
  Let~$\glie$ be a Lie~algebra.
  \begin{enumerate}
    \item
      If~$\hlie$ is another Lie~algebra and~$f \colon \glie \to \hlie$ is a homomorphism of Lie~algebras then
      \[
        f(\glie)^i
        =
        f(\glie^i)
      \]
      for every~$i \geq 0$.
    \item
      If the Lie~algebra~$\glie$ is nilpotent then any Lie~subalgebra~$\hlie$ of~$\glie$ is again nilpotent.
    \item
      If~$\glie$ is nilpotent and~$I$ is an ideal in~$\glie$ then the quotient Lie~algebra~$\glie/I$ is again nilpotent.
    \item
      If~$I$ is an ideal in~$\glie$ such that~$I$ is contained in the center~$\centerlie(\glie)$ and such that~$\glie/I$ is nilpotent then already~$\glie$ is nilpotent.
    \item
      If~$\glie$ is nonzero and nilpotent then~$\centerlie(\glie)$ is nonzero.
    \item
      If~$\glie$ is nilpotent then every~$x \in \glie$ is~{\adnilpotent}.
  \end{enumerate}
\end{proposition}


\begin{proof}
  \leavevmode
  \begin{enumerate}
    \item
      This can be shown inductively because~$f([X,Y]) = [f(X), f(Y)]$ for any two subsets~$X, Y \subseteq \glie$.
    \item
      The inclusion~$\hlie \to \glie$ is a homomorphism of Lie~algebras, hence~$\hlie^i \subseteq \glie^i$ for all~$i \geq 0$.
      It thus follows for~$n$ sufficiently large from~$\glie^i = 0$ that also~$\hlie^i = 0$.
    \item
      The canonical projection~$\pi \colon \glie \to \glie/I$ is a homorphism of Lie~algebras, hence
      \[
        (\glie/I)^i
        =
        \pi(\glie)^i
        =
        \pi(\glie^i)
      \]
      for every~$i \geq 0$.
      It thus follows for~$n$ sufficiently large from~$\glie^n = 0$ that also~$(\glie/I)^n = 0$.
    \item
      For~$n$ sufficiently large we have that~$(\glie/I)^n = 0$ and hence
      \[
        0
        =
        (\glie/I)^n
        =
        \pi(\glie)^n
        =
        \pi(\glie^n)  \,.
      \]
      This means that~$\glie^n \subseteq I \subseteq \centerlie(\glie)$.
      Then
      \[
        \glie^{n+1}
        =
        [\glie, \glie^n]
        \subseteq
        [\glie, \centerlie(\glie)]
        =
        0
      \]
      and hence~$\glie^{n+1} = 0$.
    \item
      Let~$n \geq 0$ be minimal with~$\glie^n = 0$.
      That~$\glie$ is nonzero means that~$n \geq 1$.
      Then~$\glie^{n-1} \neq 0$, but also
      \[
        [\glie, \glie^{n-1}]
        =
        \glie^n
        =
        0
      \]
      and hence~$\glie^{n-1} \subseteq \centerlie(\glie)$.
      It follows that~$\centerlie(\glie) \neq 0$.
    \item
      Let~$x \in \glie$.
      We find by induction that
      \[
        \ad(x)^n(\glie)
        \subseteq
        \glie^n
      \]
      for all~$n \geq 0$.
      It hence follows for~$n$ sufficiently large from~$\glie^n = 0$ that~$\ad(x)^n = 0$.
    \qedhere
  \end{enumerate}
\end{proof}


\begin{remark}
  The analogeous statement for \cref{solvable via ses} about nilpotency does not necessarily hold.
  Take for example the {\twodimensional} non-abelian Lie~algebra~$\glie$, that has a basis~$x$,~$y$ with~$[x,y] = x$.
  Then the {\onedimensional} linear subspace~$I \defined \kf x$ is an abelian ideal in~$\glie$ and in particular nilpotent.
  The quotient~$\glie/I$ is one-dimensional and therefore also nilpotent.
  But~$\glie$ itself is not nilpotent as seen in \cref{examples for solvable and nilpotent}.
  
  But we have the following result:
\end{remark}


\begin{corollary}
  \label{nilpotent iff nilpotent moudulo center}
  A Lie algebra~$\glie$ is nilpotent if and only if its quotient~$\glie/\centerlie(\glie)$ is nilpotent.
  \qed
\end{corollary}


\begin{remark}
  Observe that the quotient~$\glie/\centerlie(\glie)$ can be identified with the Lie~subalgebra~$\ad(\glie)$ of~$\gllie(\glie)$.
  We can therefore use \cref{solvable via ses} and \cref{nilpotent iff nilpotent moudulo center} to reduce questions about solvability or nilpotence of Lie~algebras to the case of linear Lie~algebras (observe for the application of \cref{solvable via ses} that the ideal~$\centerlie(\glie)$ is abelian and hence solvable).
\end{remark}


\begin{remark}
  If~$\glie$ is a Lie~algebra and~$I$ and~$J$ are two nilpotent ideals in~$\glie$ then it can be shown that their sum~$I+J$ is again nilpotent.
  Indeed, we have for all~$n \geq 0$ that
  \[
    (I + J)^{2n}
    \subseteq
    I^n + J^n \,.
  \]
  To see this we write
  \begin{align*}
    (I + J)^{2n}
    &=
    [I+J, [I+J, [\dotsc, [I+J, I+J] \dotsc] ] ]
    \\
    &=
    \sum_{K_i \in \{I, J\}}
    [K_1, [K_2, [\dotsc, [K_{2n}, K_{2n+1}] \dotsc] ] ] \,.
  \end{align*}
  In each summand~$[K_1, [K_2, [\dotsc, [K_{2n}, K_{2n+1}] \dotsc] ] ]$ at least~{\many{$(n+1)$}} ideals~$K_i$ are equal to the same ideal~$K \in \{I, J\}$ (the choice of which depends on the summand), say
  \[
    K_{i_1}
    =
    \dotsb
    =
    K_{i_{n+1}}
    =
    K
  \]
  with~$1_1 < i_2 < \dotsb < i_{n+1}$.
  We have that~$[L_1, L_2] \subseteq L_1, L_2$ for all ideals~$L_1, L_2 \subseteq \glie$, so we may leave out all terms~$K_i$ with~$K_i \neq K$ to get
  \begin{align*}
    {}&
    [K_1, [K_2, [\dotsc, [K_{2n}, K_{2n+1}] \dotsc] ] ]
    \\
    \subseteq{}&
    [K_{i_1}, [K_{i_n}, [\dotsc, [K_{i_n}, K_{i_{n+1}}] \dotsc] ] ]
    \\
    ={}&
    [K, [K, [\dotsc, [K, K] \dotsc] ] ]
    \\
    ={}&
    K^n
    \\
    \in{}&
    \{I^n, J^n\}  \,.
  \end{align*}
  This holds for every summand, so overall~$(I + J)^{2n} \subseteq I^n + J^n$ .
  If the ideals~$I$ and~$J$ are both nilpotent then~$I^n = J^n = 0$ for~$n$ sufficiently large, and hence~$(I + J)^{2n} = 0$.
%   WARNING: The following uses the wrong convention.
%   \begin{enumerate}
%     \item
%       For every~$n \geq 0$ we have that
%       \[
%         (I + J)^n
%         \subseteq
%         \sum_{k=0}^n I^k \cap J^{n-k}
%       \]
%       when we use the convention~$I^0 = J^0 = \glie$ (instead of~$I^0 = I$ and~$J^0 = J$).
%       This holds for~$n = 0$ because
%       \[
%         (I + J)^0
%         =
%         \glie
%         =
%         \glie \cap \glie
%         =
%         I^0 \cap J^0
%         =
%         \sum_{k=0}^0 I^k \cap J^{n-k} \,.
%       \]
%       It then follows inductively that
%       \begin{align*}
%         (I + J)^{n+1}
%         &=
%         [I + J, (I + J)^n]
%         \\
%         &\subseteq
%         \left[ I + J, \sum_{i=0}^n I^k \cap J^{n-k} \right]
%         \\
%         &=
%         \sum_{k=0}^n [I + J, I^k \cap J^{n-k}]
%         \\
%         &=
%         \sum_{k=0}^n \biggl( [I, I^k \cap J^{n-k}] + [J, I^k \cap J^{n-k}] \biggr)
%         \\
%         &\subseteq 
%         \sum_{k=0}^n \biggl( [I, I^k] \cap [I, J^{n-k}] \biggr) + \biggl( [J, I^k] \cap [J, J^{n-k}] \biggr)
%         \\
%         &=
%           \sum_{k=0}^n \biggl( [I, I^k] \cap [I, J^{n-k}] \biggr)
%         + \sum_{k=0}^n \biggl( [J, I^k] \cap [J, J^{n-k}] \biggr)
%         \\
%         &\subseteq
%           \sum_{k=0}^n \biggl( I^{k+1} \cap J^{n-k} \biggr)
%         + \sum_{k=0}^n \biggl( I^k \cap J^{n+1-k} \biggr)
%         \\
%         &\subseteq
%           \sum_{k=0}^{n+1} I^k \cap J^{n+1-k} \,.
%       \end{align*}
%     \item
%       It follows for all~$n, m \geq 0$ that
%       \[
%         (I + J)^{n+m}
%         \subseteq
%         \sum_{k=0}^n I^k \cap J^{n+m-k}
%         \subseteq
%         I^n + J^m
%       \]
%       because~$I^k \cap J^{n+m-k} \subseteq J^{n+m-k} \subseteq J^m$ whenever~$0 \leq k \leq n$ and~$I^k \cap J^{n+m-k} \subseteq I^k \subseteq I^n$ whenever~$n \leq k \leq n+m$.
%     \item
%       If the ideals~$I$ and~$J$ are both nilpotent with~$I^n = 0$ and~$J^m = 0$ then it therefore follows that~$(I + J)^{n+m} = 0$.
%   \end{enumerate}

  It follows that every finite dimensional Lie~algebra~$\glie$ admits a unique maxmial nilpotent ideal, the \defemph{nilradical}\index{nilradical} of~$\glie$.
\end{remark}





\subsection{Engel’s Theorem}


\begin{remark}
  If~$V$ is an~{\dimensional{$n$}} vector space over~$\kf$ and~$x \colon V \to V$ a nilpotent endomorphism then~$0$ is the only eigenvalue of~$x$ (and occurs with algebraic multiplicity~$n$).
  Hence there exists an eigenvector~$v \in V$ (which entails that~$v \neq 0$) with~$x(v) = 0$.
  The following proposition generalizes this observations to linear Lie~algebras consisting of nilpotent endomorphisms.
\end{remark}


\begin{proposition}
  \label{common eigenvector for nilpotent Lie algebras}
  Let~$V$ be a nonzero finite dimensional vector space and let~$\glie$ be a Lie~subalgebra of~$\gllie(V)$ consisting of nilpotent endomorphisms.
  Then there exists some nonzero~$v \in V$ with~$x(v) = 0$ for every~$x \in \glie$, i.e.~$v$ is a common eigenvector for all~$x \in \glie$ (all of which are nilpotent and thus have~$0$ as their only eigenvalue).
\end{proposition}


\begin{proof}
  We show the statement by induction over the dimension of~$\glie$.
  If~$\dim \glie = 0$ then~$\glie = 0$ and we may choose any nonzero~$v \in V$.
  If~$\dim \glie = 1$ then~$\glie = \kf x$ for any nonzero~$x \in \glie$ and we may choose any nonzero~$v \in \ker x$ (which exists because~$x$ is a nilpotent endomorphism of~$V$).
  Let now~$\dim \glie \geq 2$ and suppose that the \lcnamecref{common eigenvector for nilpotent Lie algebras} holds for all strictly smaller dimensions.
  
  We now proceed in two steps:
  In the first step we show that the \lcnamecref{common eigenvector for nilpotent Lie algebras} holds if~$\glie$ admits a nonzero proper ideal~$I$.
  We do so by applying the induction hypothesis to~$I$ and~$\glie/I$ (or rather~$\glie'$, a further quotient of~$\glie/I$).
  In the second step we show that~$\glie$ does indeed admit such an ideal.
  We do so by showing that every maximal proper Lie~subalgebra~$I$ of~$\glie$ is already an ideal, which we do by showing that~$\normallie_\glie(I) = \glie$ because~$I$ acts nilpotent on the quotient vector space~$\glie/I$.
  
  If~$\hlie$ is a proper Lie~subalgebra of~$\glie$ then the linear subspace
  \[
    W_\hlie
    \defined
    \{
      v \in V
    \suchthat
      \text{$x.v = 0$ for every~$x \in \hlie$}
    \}
  \]
  is by induction hypothesis nonzero.
  We observe that if~$I$ is an ideal in~$\glie$ then~$W_I$ is already a {\subrepresentation{$\glie$}} of~$V$:
  Indeed, we have for all~$x \in \glie$ and~$w \in W_I$ that
  \[
    z.(x.w)
    =
    x.(z.w) - [x,z].w
  \]
  for every~$z \in I$, with~$z.w = 0$ and~$[x,z].w = 0$ because~$z, [x,z] \in I$.
  
  Suppose now that~$\glie$ admits a nonzero proper ideal~$I$.
  Seen we find that~$W \defined W_I$ is a nonzero {\subrepresentation{$\glie$}} of~$V$.
  We hence get a homomorphism of Lie~algebras
  \[
    f
    \colon
    \glie
    \to
    \gllie(W) \,,
    \quad
    x
    \mapsto
    \restrict{x}{W} \,.
  \]
  The image~$\glie' \defined \im f = f(\glie)$ is a Lie~subalgebra of~$\gllie(W)$ that again consists of nilpotent endomorphisms.
  But the nonzero ideal~$I$ is contained in the kernel of~$f$, hence~$\glie'$ has strictly smaller dimension than~$\glie$.
  We hence find by induction hypothesis that there exists some nonzero~$w \in W$ with~$y.w = 0$ for every~$y \in \glie'$.
  By definition of~$\glie'$ this means that~$x.w = 0$ for every~$x \in \glie$, so we may choose~$v = w$.
  
  It remains to show that~$\glie$ admits a nonzero proper ideal.
  Let~$I$ be a proper Lie~subalgebra of~$\glie$ of maximal dimension.
  Then~$I$ is nonzero, because~$\glie$ contains nonzero proper Lie~subalgebras (e.g.~$\kf x$ with~$x \in \glie$, where we use that~$\dim \glie \geq 2$), and maximal among all Lie~subalgebras of~$\glie$ with respect to inclusion.
  We claim that~$I$ is already an ideal in~$\glie$, which is then the desired nonzero proper ideal.
  
  Indeed, for every~$x \in I$ the adjoint action~$\ad(x) = [x,-]$ maps~$I$ into itself, and hence descends to an endomorphism~$\induced{x} \colon \glie/I \to \glie/I$.
  We hence get a homomorphism of Lie~algebras
  \[
    g
    \colon
    I
    \to
    \gllie(\glie/I)  \,,
    \quad
    z
    \mapsto
    \induced{z} \,.
  \]
  The image of~$g$ is a Lie subalgebra of~$\gllie(\glie/I)$ of dimension strictly smaller than~$\glie$.
  So by induction hypothesis there exists some nonzero~$y \in \glie/I$ with~$\induced{z}.y = 0$ for every~$x \in I$.
  If~$x \in \glie$ is a representation of~$y$ then this means that~$x \notin I$ (as~$y$ is nonzero) but~$[z,x] = z.x \in I$ for every~$z \in I$ (because~$\induced{z}.y = 0$ for every~$z \in I$.
  Hence~$x$ is contained in the normalizer~$\normallie_\glie(I)$, but not contained in~$I$.
  This shows that~$I$ is properly contained in its normalizer.
  This normalizer is again a Lie~subalgebra of~$\glie$, and hence it follows from the maximality of~$\glie$ (among all proper Lie~subalgebras of~$\glie$) that~$\normallie_\glie(I) = \glie$.
  This means that~$I$ is an ideal in~$\glie$.
\end{proof}


\begin{recall}[Triangularization and invariant flags]
  \label{triangular recall}
  Let~$V$ be a finite dimensional vector space.
  \begin{itemize}
    \item
      A \defemph{flag}\index{flag} in~$V$ is an increasing sequence
      \[
        0
        =
        V_0
        \subsetneq
        V_1
        \subsetneq
        V_2
        \subsetneq
        \dotsb
        \subsetneq
        V_{n-1}
        \subsetneq
        V_n
        =
        V
      \]
      of linear subspaces~$V_i$ of~$V$.
      Note that by (our) convention every flag starts with~$0$ and ends with~$V$.
      If~$n = \dim(V)$ then the following conditions on a flag~$(V_i)_{i=0}^m$ are equivalent:
      \begin{equivalenceslist}
        \item
          $m = n$,
        \item
          $\dim V_i = i$ for every~$i = 0, \dotsc, m$,
        \item
          $(V_i)_{i=0}^m$ is a flag of maximal length,
        \item
          every quotient~$V_i/V_{i-1}$ with~$i = 1, \dotsc, m$ is {\onedimensional}.
      \end{equivalenceslist}
      The flag~$(V_i)_{i=0}^m$ is \defemph{complete}\index{complete flag}\index{flag!complete} if it satisfies these equivalent properties.
      
      Every basis~$v_1, \dotsc, v_n$ of~$V$ brings with it an \defemph{associated complete flag}~$(V_i)_{i=0}^n$\index{associated flag} of~$V$ that is given by~$V_i = \vspan_\kf(v_1, \dotsc, v_i)$ for every~$i = 0, \dotsc, n$.
      It follows by extending bases that every complete flag of~$V$ is associated to some basis of~$V$.
      
    \item
      Let~$f$ be an endomorphism of~$V$.
      A flag~$(V_i)_{i=0}^n$ is~\defemph{\invariant{$f$}}\index{invariant flag}\index{flag!invariant} if every term~$V_i$ is~{\invariant{$f$}} in the sense that~$f(V_i) \subseteq V_i$.
      
      The endomorphism~$f$ is represented by an upper triangular matrix with respect to a basis~$v_1, \dotsc, v_n$ of~$V$ if and only if the associated complete flag~$(V_i)_{i=0}^n$ is~{\invariant{$f$}}.
      It follows in particular that~$f$ is triangularizable if and only if~$V$ admits a complete~{\invariant{$f$}} flag.
      
      Observe that if the endomorphism~$f$ is nilpotent and and represented by an upper triangular matrix with respect to some basis of~$V$, then this representing matrix is nilpotent and must therefore already be strictly upper triangular.
      If~$(V_i)_{i=0}^n$ is a complete~{\invariant{$f$}} flag then it follows that already~$f(V_i) \subseteq V_{i-1}$ for every~$i \geq 1$.
    \item
      Let~$\mathcal{F}$ be any family of endomorphisms of~$V$.
      A flag~$(V_i)_{i=0}^n$ of~$V$ is~\defemph{\invariant{$\mathcal{F}$}}\index{invariant flag}\index{flag!invariant} if every term~$V_i$ is~{\invariant{$\mathcal{F}$}} in the sense that~$V_i$ is~{\invariant{$f$}} for every~$f \in \mathcal{F}$.
      Then for a basis~$v_1, \dotsc, v_n$ of~$V$ the associated complete flag~$(V_i)_{i=0}^n$ is~{\invariant{$\mathcal{F}$}} if and only if every~$f \in \mathcal{F}$ is given by an upper triangular matrix with respect to this basis.
      It follows in particular that the endomorphisms~$f \in \mathcal{F}$ are simultaneous triangularizable if and only if~$V$ admits a complete~{\invariant{$\mathcal{F}$}} flag.
  \end{itemize}
\end{recall}


\begin{lemma}[Triangularization lemma\footnote{The author is to blame for this \lcnamecref{triangularization lemma}.}]
  \label{triangularization lemma}
  Let~$\mathcal{V}$ be a class of finite dimensional representations of a Lie algebra~$\glie$ such that
  \begin{itemize}
    \item
      $\mathcal{V}$ is closed under taking quotients,
    \item
      every nonzero representation~$V \in \mathcal{V}$ admits a {\onedimensional} subrepresentation.
  \end{itemize}
  Then every representation~$V \in \mathcal{V}$ admits a complete flag of subrepresentations, and hence a basis with respect to which the action of every~$x \in \glie$ is given by an upper triangular matrix.
\end{lemma}


\begin{proof}
  We fix~$V \in \mathcal{V}$ and show the claim by induction of the dimension~$n$ of~$V$.
  For~$n = 0$ and~$n = 1$ any flag sufficies.
  If~$n \geq 2$ then~$V$ admits a {\onedimensional} subrepresentation~$U$.
  The quotient~$W \defined V/U$ is again contained in~$\mathcal{V}$ but of dimension~$n-1$ and hence admits a complete flag of subrepresentations
  \[
    0
    =
    W_0
    \subsetneq
    W_1
    \subsetneq
    W_2
    \subsetneq
    \dotsb
    \subsetneq
    W_{n-2}
    \subsetneq
    W_{n-1}
    =
    W \,.
  \]
  Let~$\pi \colon V \to V/U = W$ denote the canonical projection and for every~$i \geq 1$ let~$V_i \defined \pi^{-1}(W_{i-1})$.
  Note that~$U = V_1$, so that with~$V_0 \defined 0$ the sequence
  \[
    0
    =
    V_0
    \subsetneq
    V_1
    \subsetneq
    V_2
    \subsetneq
    \dotsb
    \subsetneq
    V_{n-1}
    \subsetneq
    V_n
    =
    V
  \]
  is a complete flag of subrepresentations for~$V$.
  This shows the first assertion.
  
  It follows from \cref{triangular recall} that any such complete flag of subrepresentations of~$V$ is associated to a basis for~$V$ with respect to which the action of every~$x \in \glie$ is given by an upper triangular matrix.
  This shows the second assertion.
\end{proof}


\begin{proposition}
  \label{characterizations of linear lie algebras consisting of nilpotent endomorphisms}
  Let~$V$ be a vector space of finite dimension~$n \defined \dim V$ and let~$\glie$ be a Lie~subalgebra of~$\gllie(V)$.
  Then the following conditions on~$\glie$ are equivalent:
  \begin{equivalenceslist}
    \item
      \label{engels g consists of nilpotent endomorphisms}
      The Lie~algebra~$\glie$ consists of nilpotent endomorphisms.
    \item
      \label{engels there exists a complete flag shifted by g}
      There exists a complete flag~$(V_i)_{i=0}^n$ of~$V$ such that~$x(V_i) \subseteq V_{i-1}$ for all~$x \in \glie$ and~$i = 1, \dotsc, n$.
    \item
      \label{engels represented by strictly upper triangular matrices}
      There exists a basis of~$V$ with respect to which every~$x \in \glie$ is represented by a strictly upper triangular matrix.
  \end{equivalenceslist}
\end{proposition}


\begin{proof}
  \leavevmode
  \begin{description}
    \item[\ref*{engels g consists of nilpotent endomorphisms}~$\implies$~\ref*{engels represented by strictly upper triangular matrices}]
      The vector space~$V$ is the natural representation of the Lie~algebra~$\glie$.
      Let~$\mathcal{V}$ be the class of all quotient representations of~$V$ and observe that~$V \in \mathcal{V}$.
      Every~$x \in \glie$ acts nilpotently on~$V$ and hence also on every quotient of~$V$.
      We can therefore apply Engel’s theorem to the image of the associated Lie~algebra homomorphism~$\rho \colon \glie \to \gllie(W)$.
      It follows that for every nonzero~$W \in \mathcal{V}$ there exists some nonzero~$x \in W$ with~$x.w = 0$ for every~$x \in \glie$;
      then~$\vspan_\kf(v)$ is a {\onedimensional} subrepresentation of~$W$.
      
      This show that we can apply \cref{triangularization lemma} ot~$\mathcal{V}$.
      There hence exists a basis of~$V$ with respect to which the action of every~$x \in \glie$ is given by an upper triangular matrix.
      Every such action is nilpotent, and hence must already be represented by a strictly triangular matrix.
    \item[\ref*{engels represented by strictly upper triangular matrices}~$\implies$~\ref*{engels there exists a complete flag shifted by g}]
      This follows from \cref{triangular recall}.
    \item[\ref*{engels there exists a complete flag shifted by g}~$\implies$~\ref*{engels g consists of nilpotent endomorphisms}]
      It follows that~$x^n = 0$.
    \qedhere
  \end{description}
\end{proof}


\begin{theorem}[Engel]
  \index{Engel’s theorem}
  A finite dimensional Lie~algebra~$\glie$ is nilpotent if and only if all its elements are~{\adnilpotent}.
\end{theorem}


\begin{proof}
  If~$\glie$ is nilpotent then every~$x \in \glie$ is~{\adnilpotent} by \cref{properties of nilpotence}.
 
  Suppose on the other hand that~$\glie$ consists of~{\adnilpotent} elemenents.
  Then~$\glie' \defined \ad(\glie)$ is a Lie subalgebra of~$\gllie(\glie)$ consisting of nilpotent endomorphisms.
  According to \cref{characterizations of linear lie algebras consisting of nilpotent endomorphisms} there exists a basis of~$V$ with respect to which every~$x \in \glie'$ is given by a strictly upper triangular matrix.
  This shows that~$\glie'$ is isomorphic to a Lie~subalgebra of some~$\nlie_n(\kf)$ (for~$n = \dim \glie$) and hence nilpotent.
  
  This shows that~$\glie/\centerlie(\glie) \cong \glie'$ is nilpotent.
  It follows by \cref{characterizations of linear lie algebras consisting of nilpotent endomorphisms} that~$\glie$ is nilpotent.
\end{proof}


\begin{remark}
  It is not true that every nilpotent Lie~subalgebra~$\glie$ of~$\gllie(V)$, where~$V$ is a finite dimensional vector space, can be represented by strictly upper triangular matrices with respect to some suitable basis of~$V$.
  An example for this is the {\onedimensional} Lie~subalgebra~$\glie \defined \kf {\id_V}$ of~$\gllie(V)$.
  This Lie algebra is abelian and hence nilpotent, but is with respect to every basis of~$V$ represented by the scalar matrices~$\kf I$.
\end{remark}


\begin{corollary}
  \label{ideals in nilpotent does intersection center}
  If~$I$ is a nonzero ideal in a finite dimensional nilpotent Lie~algebra~$\glie$ then~$I \cap \centerlie(\glie) \neq 0$.
\end{corollary}


\begin{proof}
  The Lie~algebra~$\glie$ acts on itself via the adjoint representation and~$\ad(x)$ is nilpotent for every~$x \in \glie$ by Engel’s theorem.
  The ideal~$I$ is a subrepresentation of~$\glie$ and the restriction~$\restrict{\ad(x)}{I}$ is again nilpotent for every~$x \in \glie$.
  The image of the resulting Lie~algebra homomorphism
  \[
    \glie
    \to
    \gllie(I) \,,
    \quad
    x
    \mapsto
    \restrict{\ad(x)}{I}
  \]
  does therefore consists of nilpotent endomorphisms.
  By \cref{common eigenvector for nilpotent Lie algebras} there exists some non-zero~$y \in I$ with~$\restrict{\ad(x)}{I}(y) = 0$ for every~$y \in \glie$, i.e.~$[\glie, y] = 0$.
  Then~$y \in \centerlie(\glie) \cap I$.
\end{proof}





\subsection{Lie’s Theorem}


\begin{convention}
  From now on \emph{all} occuring fields are required to be algebraically closed and of characteristic~$0$, unless otherwise specified.
  This convention holds for the rest of these notes.
  In particular all occuring Lie~algebras and vector spaces will have such a field as their ground field, even if not explicitely stated.
\end{convention}


\begin{definition}
  Let~$V$ be a representation of a Lie~algebra~$\glie$.
  For every linear functional~$\lambda \in \glie^*$ the linear subspace~$\gls*{weight space}$ of~$V$ given by
  \[
    V_\lambda
    \defined
    \{
      v \in V
    \suchthat
      \text{$x.v = \lambda(x) v$ for every~$x \in \glie$}
    \}
  \]
  is the \defemph{weight space}\index{weight!space} of~$V$ with respect to~$\lambda$.
  An element~$\lambda \in \glie^*$ with~$V_\lambda \neq 0$ is a \defemph{weight}\index{weight} of~$V$.
\end{definition}


\begin{lemma}[Invariance lemma]
  \index{invariance lemma}
  \index{lemma!invariance}
  Let~$V$ be a finite dimensional representation of a Lie~algebra~$\glie$ and let~$I$ be an ideal in~$\glie$.
  Then~$V$ is also a representation of~$I$ by restricting the action of~$\glie$ to~$I$.
  For~$\lambda \in I^*$ let~$V_\lambda$ be the~{\weightspace{$I$}} of~$V$ with respect to~$\lambda$.
  Then~$V_\lambda$ is already a~{\subrepresentation{$\glie$}} of~$V$.
\end{lemma}


\begin{proof}
  For~$v \in V$ and~$x_1, \dotsc, x_n \in \glie$ we will write
  \[
    x_1 \dotsm x_n v
    \defined
    x_1.(x_2.( \dotsc x_{n-1}.(x_n.v) \dotsc )) \,.
  \]
  The case~$V_\lambda = 0$ is welll understood, so for the rest of this proof we fix some~$\lambda \in I^*$ with~$V_\lambda \neq 0$.
 
  That~$V_\lambda$ is a {\subrepresentation{$\glie$}} of~$V$ means that~$yv \in V_\lambda$ for all~$y \in \glie$ and~$v \in V_\lambda$, which  means that~$xyv = \lambda(x)yv$ for all~$x \in I$,~$y \in \glie$ and~$v \in V_\lambda$.
  We see that
  \[
    xyv
    =
    [x,y]v + yxv
    =
    \lambda([x,y])v + \lambda(x)yv
  \]
  for all~$x \in I$,~$y \in \glie$ and~$v \in V_\lambda$.
  We hence need to show that~$\lambda([x,y]) = 0$ for all~$x \in I$ and~$y \in \glie$.
 
  Until further notice we fix some~$y \in \glie$ and some nonzero~$v \in V_\lambda$.
  As~$V$ is finite dimensional there exists some maximal~$n \geq 1$ such that the vectors~$v, yv, \dotsc, y^n v$ are linearly independent.
  Let
  \[
    W_i
    \defined
    \vspan_k(v, yv, \dotsc, y^i v)
  \]
  for every~$i = 0, \dotsc, n$.
  The linear space~$W_n$ is invariant under the action of~$y$ because the vector~$y^{n+1} v$ is linearly dependent on the vectors~$v, \dotsc, y^n v$.
 
  \begin{claim*}
    The linear subspaces~$W_1, \dotsc, W_n$ are~{\subrepresentations{$I$}} of~$V$.
    With respect to the basis~$w, y w, \dotsc, y^i w$ of~$W_i$ the action of any~$x \in I$ is represented by an upper triangular matrix whose every diagonal entry is~$\lambda(x)$.
  \end{claim*}

 \begin{proof}[Proof of the claim]
    We prove the claim by induction over~$i$.
    The claim holds for~$i = 0$ since~$W_0 = \kf v$ is spanned by the vector~$v$ which is a~{\weightvector{$\lambda$}} for the action of~$I$.
    Suppose now that~$i < n$ and that the claim holds for the linear subspaces~$W_0, \dotsc, W_i$.
    If~$x \in I$ then also~$[x,y] \in I$ and therefore
    \[
      x y^{i+1} v
      =
      \underbrace{[x,y] y^i v}_{\mathclap{\substack{\in W_i \\ \text{by induction}}}} + y x y^i v
      \equiv
      y x y^i v
      \mod
      W_i \,.
    \]
    We find by the induction hypothesis that~$x y^i v \in W_i$ because~$W_i$ is a subrepresentation, and also that this vector is of the form
    \[
      x y^i v
      \equiv
      \lambda(x) y^i v
      \mod
      W_{i-1} \,.
    \]
    It follows that
    \[
      y x y^i v
      \equiv
      \lambda(x) y^{i+1} v
      \mod W_i  \,,
    \]
    because~$y W_{i-1} \subseteq W_i$.
    This shows the claim for~$W_{i+1}$.
 \end{proof}
  
  Let~$x \in I$.
  It follows from the above claim that the~{\dimensional{$(n+1)$}} linear subspace~$W_n$ is invariant under the action of~$[x,y]$ because~$[x,y] \in I$.
  This action is given by an endomorphism~$\phi_{[x,y]} \in W_n \to W_n$, that is represented with respect to some suitable basis of~$W_n$ by an upper triangular matrix whose every diagonal entry is~$\lambda([x,y])$.
  It follows that in particular
  \begin{equation}
    \label{invariance lemma zero trace}
    \tr \phi_{[x,y]}
    =
    (n+1) \lambda([x,y])  \,.
  \end{equation}
  On the other hand the linear subspace~$W_n$ is invariant under the action of both~$x$ (by the claim because~$x \in I$) and under the action of~$y$ (as seen above before the claim).
  These two elements act by endomorphisms~$\phi_x$ and~$\phi_y$ on~$W_n$.
  Because~$V$ is a representation of the Lie~algebra~$\glie$ it follows that~$\phi_{[x,y]} = [\phi_x, \phi_y]$ and thus~$\tr \phi_{[x,y]} = 0$.
  Together with \eqref{invariance lemma zero trace} it follows that~$\lambda([x,y]) = 0$.
\end{proof}


\begin{theorem}[Lie]
  \index{Lie’s theorem}
  Let~$\glie$ be a solvable Lie~subalgebra of~$\gllie(V)$ for some nonzero finite dimensional~{\vectorspace{$\kf$}}. Then there exists a common eigenvector for~$\glie$, i.e.\ some nonzero~$v \in V$ with~$x(v) \in \kf v$ for every~$x \in \glie$.
\end{theorem}


\begin{proof}
  We show the statement by induction over the dimension~$n$ of~$\glie$.
  If~$n = 0$ then~$\glie = 0$ and any nonzero~$v \in V$ does the job.
  If~$n = 1$ then~$\glie = \kf x$ for some nonzero~$x \in \gllie(V)$.
  Then any eigenvector of~$x$ does the job (and such an eigenvector exists because the field~$\kf$ is assumed to be algebraically closed and~$V \neq 0$).
 
  Suppose that~$n \geq 2$ and that the statement holds for every smaller dimension.
  The proof proceeds in four steps:
  \begin{itemize}
    \item
      Finding an ideal~$I$ in~$\glie$ of codimension~$1$.
    \item
      Finding a common eigenvectors for~$I$ by induction hypothesis.
    \item
      Showing that~$\glie$ stabilizes as nonzero subspace~$U \subseteq V$ of such eigenvectors.
    \item
      Writing~$\glie = I \oplus \kf y$ (as vector spaces) and finding an eigenvector of~$y$ in~$U$.
  \end{itemize}

  For the first step we note that~$[\glie,\glie]$ in~$\glie$ because~$\glie$ is nonzero and solvable.
  Hence~$\glie/[\glie,\glie]$ is a nonzero abelian Lie~algebra.
  Any linear subspace~$J \subseteq \glie/[\glie,\glie]$ of codimension~$1$ is then an ideal in~$\glie/[\glie,\glie]$.
  Hence the preimage~$I = \pi^{-1}(J)$ under the canonical projection~$\glie \to \glie/[\glie,\glie]$ is an ideal in~$\glie$ of codimension~$1$.
 
  For the second step we note~$I$ is solvable because it is a Lie~subalgebra of~$\glie$.
  So by induction hypothesis there exists a common eigenvector for~$I$.
  Hence there exists some~$\lambda \in I^*$ with~$U \defined V_\lambda \neq 0$, where~$V_\lambda$ denotes the~{\weightspace{$I$}} with respect to~$\lambda$.
  
  For the third step we apply the invariance lemma to see that~$U$ is a~{\subrepresentation{$\glie$}} of~$V$.
 
  For the fourth step we may write~$\glie = I \oplus \kf y$ for some~$y \in \glie$ because~$I$ has codimension~$1$.
  Then~$y$ stabilizs~$U$ and hence admits an eigenvector~$v \in U$, as~$\kf$ is algebraically closed.
  This vector~$v$ is then a common eigenvector for~$\glie$.
\end{proof}


\begin{remark}
 The idea of organizing the proof into four steps is taken from~\cite[\S 4.1]{Humphreys}.
%  This organizational structure also serves to emphasize the similarities with the proof of Proposition~\ref{common eigenvector for nilpotent Lie algebras}.
\end{remark}


\begin{remark}
  Observe that for a Lie~subalgebra~$\glie$ of some~$\gllie(V)$, where~$V$ is some vector space, a vector~$v \in V$ is a common eigenvector for~$\glie$ if and only if the the subspace~$\vspan_\kf(v)$ is a {\onedimensional}~{\subrepresentation{$\glie$}} of~$V$.
\end{remark}


\begin{proposition}
  \label{triangularization of solvable linear lie algebras}
  Let~$\glie$ be a Lie~subalgebra of~$\gllie(V)$ where~$V$ is a finite dimensional~{\vectorspace{$\kf$}}.
  Then the following conditions on~$\glie$ are equivalent:
  \begin{equivalenceslist}
    \item
      \label{is solvable}
      $\glie$ is solvable.
    \item
      \label{stabilizes a complete flag}
      $\glie$ stabilizes some complete flag of~$V$, i.e.\ there exists a complete flag~$(V_i)_{i=0}^n$ of~$V$~with~$x(V_i) \subseteq V_i$ for all~$x \in \glie$ and~$i = 0, \dotsc, n$.
    \item
      \label{is triangularizable}
      There exists a basis of~$V$ with respect to which every~$x \in \glie$ is represented by an upper triangular matrix.
      In particular,~$\glie$ is isomorphic to a Lie~subalgebra of~$\tlie_n(\kf)$ for~$n = \dim V$.
  \end{equivalenceslist}
\end{proposition}


\begin{proof}
  \leavevmode
  \begin{description}
    \item[\ref*{is solvable}~$\implies$~\ref*{is triangularizable}]
      The vector space~$V$ is the natural representation of the Lie~algebra~$\glie$.
      Let~$\mathcal{V}$ be the class of all quotient representations of~$V$.
      For every~$W \in \mathcal{V}$ the image of the Lie~algebra homomorphism~$\rho_W \colon\glie \to \gllie(W)$,~$x \mapsto \restrict{x}{W}$ is again solvable.
      It therefore follows from Lie’s theorem that there exists for every nonzero~$W \in \mathcal{V}$ some nonzero~$w \in W$ such that~$w$ is a common eigenvector vor~$\rho_W(\glie)$;
      then~$\vspan_\kf(v)$ is a {\onedimensional} subrepresentation of~$W$.
      
      This shows that we can apply \cref{triangularization lemma} to get the desired result.
    \item[\ref*{is triangularizable}~$\implies$~\ref*{is solvable}]
      Every Lie~subalgebra of~$\tlie_n(\kf)$ is again solvable.
    \item[\ref*{is triangularizable}~$\iff$~\ref*{stabilizes a complete flag}]
      This follows from \cref{triangular recall}.
    \qedhere
  \end{description}
\end{proof}


\begin{corollary}
  \label{solvable linear lie algebra has commutator consisting of nilpotent endomorphisms}
  If~$\glie$ is a solvable Lie~subalgebra of~$\gllie(V)$ for some finite dimensional vector space~$V$ then~$[\glie, \glie]$ consists of nilpotent endomorphisms.
\end{corollary}


\begin{proof}
  There exists some basis of~$V$ with respect to which every~$x \in \glie$ is represented by an upper triangular matrix.
  Every~$x \in [\glie, \glie]$ is then represented by a strictly upper triangular matrix and hence nilpotent.
\end{proof}


\begin{corollary}
  A finite dimensional Lie~algebra~$\glie$ is solvable if and only if~$[\glie, \glie]$ is nilpotent.
\end{corollary}


\begin{proof}
  If~$[\glie, \glie] = \glie^{(1)}$ is nilpotent then it is solvable, and hence~$\glie$ is solvable.
  
  For the converse we observe that~$\glie' \defined \ad(\glie)$ is a Lie~subalgebra of~$\gllie(\glie)$ that is again solvable.
  It follows that~$[\glie, \glie]$ consists of~{\adnilpotent} elements because
  \[
    \ad([\glie, \glie])
    =
    [\ad(\glie), \ad(\glie)]
    =
    [\glie', \glie']
  \]
  (since~$\ad$ is a homomorphism of Lie~algebras) consists of nilpotent endomorphisms of~$\glie$ by \cref{solvable linear lie algebra has commutator consisting of nilpotent endomorphisms}.
  The Lie~algebra~$[\glie, \glie]$ is hence nilpotent by Engel’s theorem.
\end{proof}


\begin{corollary}
  \label{characterization of fd solvable Lie algebras}
  For a finite dimensional Lie~algebra~$\glie$ the following conditions are equivalent:
  \begin{equivalenceslist}
    \item
      \label{g is solvable}
      The Lie~algebra~$\glie$ is solvable.
    \item
      \label{every fd rep is triangularizable}
      Every finite dimensional representation of~$\glie$ admits a basis with respect to which the action of every~$x \in \glie$ is given by an upper triangular matrix.
    \item
      \label{every fd rep admits a complete flag of subreps}
      Every finite dimensional representation of~$\glie$ admits a complete flag~$(V_i)_{i=0}^n$ consisting of subrepresentations.
    \item
      \label{every fd rep contains a one dimensional subrep}
      Every nonzero finite dimensional representation of~$\glie$ contains a {\onedimensional} subrepresentation.
    \item
      \label{every fd irrep is one dimensional}
      Every finite dimensional irreducible representation of~$\glie$ is {\onedimensional}.
  \end{equivalenceslist}
\end{corollary}


\begin{proof}
  \leavevmode
  \begin{description}
    \item[\ref*{g is solvable}~$\implies$~\ref*{every fd rep is triangularizable}]
      If~$(V, \rho)$ is any finite dimensional representation of~$\glie$ then~$\glie' \defined \rho(\glie)$ is a solvable Lie~subalgebra of~$\gllie(V)$.
      It follows from \cref{triangularization of solvable linear lie algebras} that there exists a basis of~$V$ with respect to which every~$x' \in \glie'$ is given by an upper triangular matrix.
      This means that the action of every~$x \in \glie'$ on~$V$ is given by an upper triangular matrix with respect to this basis.
    \item[\ref*{every fd rep is triangularizable}~$\iff~$\ref*{every fd rep admits a complete flag of subreps}]
      If~$(V, \rho)$ is any finite dimensional representation of~$\glie$ then~$\glie' \defined \rho(\glie)$ is a Lie subalgebra of~$\gllie(V)$.
      The equivalence follows by applying \cref{triangularization of solvable linear lie algebras} to~$\glie'$.
    \item[\ref*{every fd rep admits a complete flag of subreps}~$\implies$~\ref*{every fd rep contains a one dimensional subrep}]
      Every complete flag of subrepresentations contains a {\onedimensional} term.
    \item[\ref*{every fd rep contains a one dimensional subrep}~$\implies$~\ref*{every fd irrep is one dimensional}]
      Every finite dimensional irreducible representation~$V$ contains a {\onedimensional} subrepresentation~$U$, for which necessarily~$V = U$ by the irreduciblity of~$V$.
    \item[\ref*{every fd irrep is one dimensional}~$\implies$~\ref*{every fd rep contains a one dimensional subrep}]
      Every finite dimensional representation contains an irreducible subrepresentation (e.g.\ a subrepresentation of minimal nonzero dimension) that is then {\onedimensional}.
    \item[\ref*{every fd rep contains a one dimensional subrep}~$\implies$~\ref*{every fd rep is triangularizable}]
      We can apply \cref{triangularization lemma} to the class of all finite dimensional representations of~$\glie$.
    \item[\ref*{every fd rep is triangularizable}~$\implies$~\ref*{g is solvable}]
      We find that for the adjoint representation of~$\glie$ there exists a basis of~$\glie$ with respect to which the action of every~$x \in \glie$ (i.e.~$[x,-]$) is given by an upper triangular matrix.
      This means by \cref{triangularization of solvable linear lie algebras} that the Lie~subalgebra~$\glie' \defined \ad(\glie)$ of~$\gllie(V)$ is solvable.
      Therefore~$\glie$ is solvable because~$\glie' \cong \glie/\centerlie(\glie)$ with~$\centerlie(\glie)$ being an abelian and hence solvable ideal in~$\glie$.
    \qedhere
  \end{description}
\end{proof}


\begin{remark}
  We used the finite dimensionality of~$\glie$ only for the implication~\ref*{every fd rep is triangularizable}~$\implies$~\ref*{g is solvable}.
  The author does not know what happens for this implication when~$\glie$ is infinite dimensional.
  (He suspects that it is false, but does not know a counterexample.)
\end{remark}


\begin{remark}
  \Cref{characterization of fd solvable Lie algebras} does not hold for general fields~$\kf$, even if algebraically closed.
  To see this let~$\kf$ be an algebraically closed field with~$\ringchar \kf = 2$ and let~$\glie \defined \sllie_2(\kf)$.
  Recall that the standard basis of~$\glie$ is given by the matrices
  \begin{gather*}
    e
    =
    \begin{pmatrix}
      0 & 1 \\
      0 & 0
    \end{pmatrix} \,,
    \qquad
    h
    =
    \begin{pmatrix*}[r]
      1 &  0  \\
      0 & -1
    \end{pmatrix*},
    \qquad
    f
    =
    \begin{pmatrix}
      0 & 0 \\
      1 & 0
    \end{pmatrix}
  \shortintertext{with}
    [h,e] = 0 \,,
    \qquad
    [h,f] = 0 \,,
    \qquad
    [e,f] = h \,.
  \end{gather*}
  We find that the Lie algebra~$\glie$ is nilpotent (with~$\glie^2 = 0$) and in particular also solvable.
  Let~$V \defined \kf^2$ be the natural representation of~$\glie$, i.e.~$\glie$ acts on~$V$ by~$x.v = x(v)$ for all~$x \in \glie$ and~$v \in V$.
  Then
  \[
    e.\vect{x \\ y}
    =
    \vect{y \\ 0}
    \qquad\text{and}\qquad
    f.\vect{x \\ y}
    =
    \vect{0 \\ x} \,.
  \]
  It follows that if~$U \subseteq V$ is a nonzero subrepresentation then~$U$ contains one of the standard basis vectors~$e_1$,~$e_2$ of~$V$, and then also the other one.
  Hence~$U = V$, which shows that~$V$ is an irreducible representation of~$\glie$.
  But~$V$ is not {\onedimensional}.
\end{remark}
